---
title: "Preprocessing script"
format: html
editor: visual
---

# Setup

```{r}
rm(list = ls())
```

Load libraries

```{r libraries}
library(dplyr)
library(lubridate)
library(stringr)
```

Load data

```{r data}
ukb <- readRDS("/rds/general/project/hda_25-26/live/TDS/TDS_Group8/extraction_and_recoding/outputs/recoded.rds")
cvd_event <- readRDS("/rds/general/project/hda_25-26/live/TDS/TDS_Group8/Data/cvd_events.rds")
```

# Addition of Eid, CVD_event and CVD_date columns to data

```{r}
# setwd

setwd("/rds/general/project/hda_25-26/live/TDS/TDS_Group8/final_recode")

# add eid column
ukb$eid <- rownames(ukb)
rownames(ukb) <- NULL
ukb$eid <- as.character(ukb$eid)

cvd_event$eid <- as.character(cvd_event$eid)

# cvd event flag
cvd_ids <- unique(cvd_event$eid)
ukb$cvd_event <- as.integer(ukb$eid %in% cvd_ids)

# first cvd date 
first_cvd <- aggregate(date ~ eid, cvd_event, min)

ukb$cvd_date <- first_cvd$date[match(ukb$eid, first_cvd$eid)]

# reorder columns
ukb <- ukb[, c("eid", "cvd_event", "cvd_date",
               setdiff(names(ukb), c("eid", "cvd_event", "date")))]

# check binary
table(ukb$cvd_event, useNA = "ifany")

```

# Converting IDK to NAs

```{r}
#change all IDK to NAs
ukb_v2 <- ukb
ukb_v2[] <- lapply(ukb_v2, function(x) {
  if (is.character(x) || is.factor(x)) {
    x[x == "Do not know"] <- NA
  }
  x
})

#check
sum(sapply(ukb_v2, function(x) {
  if (is.factor(x)) x <- as.character(x)
  if (is.character(x)) sum(x == "Do not know", na.rm = TRUE) else 0
}))


```

# Dealing with "Prefer not to answer"

```{r}
#'prefer not to answer' rate calculation
pna_rate_nonNA <- sapply(ukb_v2, function(x) {
  if (is.character(x) || is.factor(x)) {
    mean(as.character(x) == "Prefer not to answer", na.rm = TRUE)  
  } else NA_real_
})
pna_nonNA <- data.frame(
  column = names(pna_rate_nonNA),
  pna_rate = as.numeric(pna_rate_nonNA),
  row.names = NULL
)

N <- nrow(ukb)
pna_rate <- sapply(ukb, function(x) {
  if (is.character(x) || is.factor(x)) {
    sum(as.character(x) == "Prefer not to answer", na.rm = TRUE) / N
  } else {
    NA_real_
  }
})
pna_df <- data.frame(
  column = names(pna_rate),
  pna_rate = as.numeric(pna_rate),
  row.names = NULL
)
pna_df <- pna_df[order(-pna_df$pna_rate), ]

merged_pna <- merge(
  pna_df,
  pna_nonNA,
  by = "column",
  all = TRUE,
  suffixes = c("_totalN", "_nonNA")
)

names(merged_pna)[names(merged_pna) == "pna_rate_totalN"] <- "pna_rate_overallN"
names(merged_pna)[names(merged_pna) == "pna_rate_nonNA"]  <- "pna_rate_among_nonmissing"


#change pna to na in continuous variables
char_cols <- names(ukb_v2)[sapply(ukb_v2, is.character)]

#change character to numeric
ukb_v2[char_cols] <- lapply(ukb_v2[char_cols], function(x) {
  x[x == "Prefer not to answer"] <- NA
  suppressWarnings(as.numeric(x))
})

#final check
types_df <- data.frame(
  variable = names(ukb_v2),
  class = sapply(ukb_v2, function(x) paste(class(x), collapse = "/")),
  stringsAsFactors = FALSE
)
```

# Exclusion

Based on research question will either remove or keep prevalent cases

```{r before baseline}

#cvd before baseline
ukb_v2 <- ukb_v2 %>%
  mutate(cvd_case = ifelse(!is.na(cvd_date) & cvd_date < date_recr.0.0, "prevalent", "incident"))

ukb_v2 <- ukb_v2 %>%
  filter(cvd_case != "prevalent")%>%
  select(-cvd_case)
        
```

# Reclassify ethnicity

```{r}
eth <- as.character(ukb_v2$ethnicity.0.0)
eth <- trimws(eth)
pna_set <- c("Prefer not to answer")

white_set <- c(
  "White",
  "British", "Irish", "Any other white background"
)

asian_set <- c(
  "Asian or Asian British",
  "Indian", "Pakistani", "Bangladeshi",
  "Any other Asian background",
  "Chinese"
)

mixed_set <- c(
  "Mixed",
  "White and Black Caribbean", "White and Black African", "White and Asian",
  "Any other mixed background"
)

black_set <- c("Black or Black British", "Caribbean", "African", "Any other Black background")

other_set <- c('Other ethnic group')


ukb_v2$ethnicity.0.0 <- dplyr::case_when(
  is.na(eth) ~ NA_character_,
  eth %in% pna_set ~ "Prefer not to answer",
  eth %in% white_set ~ "White",
  eth %in% asian_set ~ "Asian",
  eth %in% mixed_set ~ "Mixed",
  eth %in% black_set ~ "Black",
  eth %in% other_set ~ "Other")

ukb_v2$ethnicity.0.0  <- as.factor(ukb_v2$ethnicity.0.0)
```

# Remove outliers

```{r}
ukb_v2 <- ukb_v2 %>%
  mutate(across(
    where(is.numeric),
    ~ ifelse(. < -1e5, NA, .)
  ))

```

# Drop columns

Drop cols, keep final instance, average cols

```{r drop columns}
# only have one date of death col (both cols are the exact same)

ukb_v2$date_death.0.0.1 <- NULL 

# keep latest ses_current_employment_status

ukb_v2 <- ukb_v2[, !grepl("ses_current_employment_status\\.0\\.[0-5]$", names(ukb_v2))]

# keep the latest ses_commute_transport_type

ukb_v2 <- ukb_v2[, !grepl("ses_commute_transport_type\\.0\\.[0-2]$", names(ukb_v2))]

# dropping employment_chemical_fume, employment_cigarette_smoke and taking the latest instance for employment_noise

ukb_v2 <- ukb_v2 %>%
  mutate(employment_noise = employment_noise.0.39)%>%
  select(-matches("^employment_noise\\.0\\."), -matches("^employment_chemical_fume\\.0\\."),
         -matches("^employment_cigarette_smoke\\.0\\."))

# take the latest qualification col

ukb_v2 <- ukb_v2[, !grepl("qualification\\.0\\.[0-4]$", names(ukb_v2))]

# dropping vars

ukb_v2 <- ukb_v2%>%
  select(-env_domestic_garden_pct_buffer_1000m.0.0, -env_greenspace_pct_buffer_1000m.0.0, -env_natural_env_perc_1000.0.0, -env_no2_air_pollution_2005.0.0, -env_no2_air_pollution_2006.0.0, -env_no2_air_pollution_2007.0.0, -env_water_perc_1000.0.0, -sbp_automated.0.0, -sbp_automated.0.1, -yob.0.0, -mob.0.0, -cvd_date.1, -mh_irritability.0.0, -mh_loneliness_isolation.0.0, -mh_miserableness.0.0,-mh_nervous_feelings.0.0, -mh_mood_swings.0.0, -mh_sensitivity.0.0, -mh_suffer_nerves.0.0, -mh_tense.0.0, -mh_worrier.0.0, -mh_worry_embarrassment.0.0, -mh_fed_feelings.0.0, -mh_guilty_feelings.0.0, -phys_walk_duration.0.0, -phys_vigorous_activity_duration.0.0, -phys_usual_walking_pace.0.0, -phys_strenuous_sports_freq_last4w.0.0, -phys_strenuous_sports_duration.0.0, -phys_mvpa_walking_rec_met.0.0, -phys_mvpa_rec_met.0.0, -phys_moderate_activity_duration.0.0, -phys_light_diy_freq_last4w.0.0, -phys_light_diy_duration.0.0, -phys_days_per_week_walked_10plus_min.0.0, -phys_days_per_week_vigorous_activity_10plus_min.0.0, -phys_days_per_week_moderate_activity_10plus_min.0.0, -phys_heavy_diy_freq_last4w.0.0, -phys_heavy_diy_duration.0.0, -alc_monthly_beer_cider.0.0, -alc_monthly_champagne_white_wine.0.0, -alc_monthly_fortified_wine.0.0, -alc_monthly_other_alcohol_drink.0.0, -alc_monthly_red_wine.0.0, -alc_monthly_spirits.0.0, -alc_weekly_beer_cider.0.0, -alc_weekly_champagne_white_wine.0.0, -alc_weekly_fortified_wine.0.0, -alc_weekly_other_alc_drinks.0.0, -alc_weekly_red_wine.0.0, -alc_weekly_spirits.0.0, -alc_drinker_former.0.0, -alc_amount_daily.0.0, -alc_six_more_frequency.0.0, -phys_walking_for_pleasure_duration.0.0, -phys_walking_for_pleasure_freq_last4w.0.0, -smoke_maternal_around_birth.0.0,-smoke_age_start_current.0.0, -smoke_age_start_former_smoker.0.0, -smoke_age_stop.0.0,
-smoke_age_stopped_cig.0.0, -smoke_people_in_household.0.0, -smoke_tobacco_exposure_home.0.0,-smoke_tobacco_exposure_outside.0.0)


# average dbp_automated and dbp_manual cols

ukb_v2 <- ukb_v2 %>%
  mutate(dbp_avg = rowMeans(select(., dbp_automated.0.0, dbp_automated.0.1, dbp_manual.0.0, dbp_manual.0.1),na.rm = TRUE)) %>%
  select(-dbp_automated.0.0, -dbp_automated.0.1, -dbp_manual.0.0, -dbp_manual.0.1)

# average sbp_manuals cols

ukb_v2 <- ukb_v2%>%
  mutate(sbp_avg = rowMeans(select(., sbp_manual.0.0, sbp_manual.0.1), na.rm = TRUE))%>%
  select(-sbp_manual.0.0, -sbp_manual.0.1)

# average fev and fvc as multiple blows are done so should take an average 

ukb_v2 <- ukb_v2%>%
  mutate(fev_avg = rowMeans(select(., fev1.0.0, fev1.0.1, fev1.0.2), na.rm = TRUE))%>%
  mutate(fvc_avg = rowMeans(select(., fvc.0.0, fvc.0.1, fvc.0.2), na.rm = TRUE))%>%
  select(-fvc.0.0, -fvc.0.1, -fvc.0.2, -fev1.0.0, -fev1.0.1, -fev1.0.2)

# creating mob and yob from dob

ukb_v2$mob <- month(ukb_v2$dob.0.0, label = TRUE)
ukb_v2$yob <- year(ukb_v2$dob.0.0)

```

# Compute missing rate and drop variables

```{r}
# column-level missingness
missing_rate <- sapply(ukb_v2, function(x) mean(is.na(x)))

missing_summary <- data.frame(
  var = names(ukb_v2),
  missing_rate = missing_rate
)

# find variables whose missing rate > 50% 
vars_to_drop <- names(missing_rate[missing_rate > 0.5])

vars_to_keep_even_if_high_missing <- c("cvd_date", "date_death.0.0")

vars_to_drop <- setdiff(vars_to_drop, vars_to_keep_even_if_high_missing)

# drop
ukb_v2 <- ukb_v2[, !(names(ukb_v2) %in% vars_to_drop)]

```

# Converting PNA to NA for scores

```{r PNA to NA}
ukb_v3 <- ukb_v2

cols_to_clean <- c(
  "diet_spread_type.0.0",
  "diet_salt_added_food.0.0",
  "diet_milk_type.0.0",
  "diet_processed_meat_intake.0.0",
  "diet_pork_intake.0.0",
  "diet_oily_fish_intake.0.0",
  "diet_non_oily_fish_intake.0.0",
  "diet_lamb_mutton_intake.0.0",
  "diet_beef_intake.0.0",
  "med_chol_bp_diabetes.0.0",
  "med_chol_bp_diabetes.0.1",
  "med_chol_bp_diabetes.0.2",
  "med_chol_bp_diabetes_hormones.0.0",
  "med_chol_bp_diabetes_hormones.0.1",
  "med_chol_bp_diabetes_hormones.0.2",
  "med_chol_bp_diabetes_hormones.0.3",
  "med_pain_relief_constipation_heartburn.0.0",
  "med_pain_relief_constipation_heartburn.0.1",
  "med_pain_relief_constipation_heartburn.0.2",
  "med_pain_relief_constipation_heartburn.0.3",
  "med_pain_relief_constipation_heartburn.0.4",
  "med_pain_relief_constipation_heartburn.0.5",
  "sleep_daytime_dozing.0.0",
  "sleep_insomnia.0.0",
  "sleep_snore.0.0",
  "sleep_chronotype.0.0"
)

ukb_v4 <- ukb_v3 %>%
  mutate(across(all_of(cols_to_clean),
                ~ na_if(.x, "Prefer not to answer")))
```

# Logic Skipping - smoke

```{r}
current_smoker <- ukb_v4$smoke_tobacco_current.0.0 =="Yes, on most or all days"
not_current_smoker <- !current_smoker
ukb_v4$smoke_current_indicator <- ifelse(current_smoker, 1, 0)

ukb_v4$smoke_tobacco_type_current.0.0 <- as.character(ukb_v4$smoke_tobacco_type_current.0.0)
ukb_v4$smoke_tobacco_type_current.0.0[not_current_smoker] <- "None"
ukb_v4$smoke_number_daily_current_smoker.0.0[not_current_smoker] <- 0


# field 1249
past_smoker <- ukb_v4$smoke_tobacco_past.0.0 == 'Smoked on most or all days'
not_past_smoker <- !past_smoker
ukb_v4$smoke_past_indicator <- ifelse(past_smoker, 1, 0)

ukb_v4$smoke_tobacco_type_previous.0.0 <- as.character(ukb_v4$smoke_tobacco_type_previous.0.0)
ukb_v4$smoke_tobacco_type_previous.0.0[not_past_smoker] <- 'None'
ukb_v4$smoke_cig_daily_previous.0.0[not_past_smoker] <- 0


# field 3446
cigar_pipe <- ukb_v4$smoke_tobacco_type_current.0.0 == "Cigars or pipes"
not_cigar_pipe <- !cigar_pipe 
ukb_v4$cigar_pipe_indicator <- ifelse(cigar_pipe, 1, 0)
ukb_v4$smoke_number_cig_previous_current_cigar_pipe.0.0[not_cigar_pipe] <- 0
```

# Logic Skipping - diet

```{r}
##coffee-------------------
#add "None" as answer to "diet_coffee_type.0.0" if "diet_coffee_intake" = 0
ukb_v4$diet_coffee_type.0.0 <- as.character(ukb_v4$diet_coffee_type.0.0)
ukb_v4$diet_coffee_type.0.0[ukb_v4$diet_coffee_intake.0.0 == 0] <- "None"

##bread----------------------------
#add "None" as answer to "diet_bread_type.0.0" if "diet_bread_intake" = 0
ukb_v4$diet_bread_type.0.0 <- as.character(ukb_v4$diet_bread_type.0.0)
ukb_v4$diet_bread_type.0.0[ukb_v4$diet_bread_intake.0.0 == 0] <- "None"

#cereal----------------------------
#add "None" as answer to "diet_cereal_type.0.0" if "diet_cereal_intake" = 0
ukb_v4$diet_cereal_type.0.0 <- as.character(ukb_v4$diet_cereal_type.0.0)
ukb_v4$diet_cereal_type.0.0[ukb_v4$diet_cereal_intake.0.0 == 0] <- "None"

#alc_wih_meals---------------------
#add "None" as answer to "alc_with_meals.0.0" if "alc_freq" = "Never"
ukb_v4$alc_freq.0.0 <- as.character(ukb_v4$alc_freq.0.0)
ukb_v4$alc_with_meals.0.0 <- as.character(ukb_v4$alc_with_meals.0.0)
ukb_v4$alc_with_meals.0.0[ukb_v4$alc_freq.0.0 == "Never"] <- "None"
```

# Logic Skipping - med

```{r}
# med_chol_bp_diabetes.0\.\0-2 <- male
# med_chol_bp_diabetes_hormones.0\.\1-3 <- female

# base off sex.0.0

med_male <- grep("^med_chol_bp_diabetes\\.0\\.[0-2]+$",
                             names(ukb_v4),
                             value = TRUE)

med_female <- grep("^med_chol_bp_diabetes_hormones\\.0\\.[0-3]+$",
                               names(ukb_v4),
                               value = TRUE)

# Male meds: convert to character first, then apply NA for females
ukb_v4 <- ukb_v4 %>%
  mutate(across(all_of(med_male),
                ~ ifelse(sex.0.0 == "Female", NA, as.character(.)))) %>%
  
  # Female meds: convert to character first, then apply NA for males
  mutate(across(all_of(med_female),
                ~ ifelse(sex.0.0 == "Male", NA, as.character(.))))
```

# Logic Skipping - ses

```{r}
# create employed indicator 

ukb_v4 <- ukb_v4 %>%
  mutate(
    employed = ifelse(
      ses_current_employment_status.0.6 %in%
        c("In paid employment or self-employed"),
      1, 0
    )
  )

#creating chars
ukb_v4$ses_job_shift_work.0.0 <- as.character(ukb_v4$ses_job_shift_work.0.0)
ukb_v4$ses_job_mainly_walking_or_standing.0.0 <- as.character(ukb_v4$ses_job_mainly_walking_or_standing.0.0)
ukb_v4$ses_job_heavy_manual_or_physical_work.0.0 <- as.character(ukb_v4$ses_job_heavy_manual_or_physical_work.0.0)
ukb_v4$ses_commute_transport_type.0.3 <- as.character(ukb_v4$ses_commute_transport_type.0.3)

# ses_length_working_week.0.0
ukb_v4$ses_length_working_week.0.0[ukb_v4$employed %in% c(0)] <- 0

# ses_job_shift_work.0.0
ukb_v4$ses_job_shift_work.0.0[ukb_v4$employed %in% c(0)] <- NA

# ses_job_mainly_walking_or_standing.0.0
ukb_v4$ses_job_mainly_walking_or_standing.0.0[ukb_v4$employed %in% c(0)] <- NA

# ses_job_heavy_manual_or_physical_work
ukb_v4$ses_job_heavy_manual_or_physical_work.0.0[ukb_v4$employed %in% c(0)] <- NA

# ses_commute_transport_type
ukb_v4$ses_commute_transport_type.0.3[ukb_v4$employed %in% c(0)] <- NA

# ses_commute_frequency
ukb_v4$ses_commute_frequency.0.0[ukb_v4$employed %in% c(0)] <- 0

# ses_commute_distance_home_to_work
ukb_v4$ses_commute_distance_home_to_work.0.0[ukb_v4$employed %in% c(0)] <- 0

# ses_age_completed_full_time_education
ukb_v4$ses_age_completed_full_time_education.0.0[ukb_v4$qualification.0.5 %in% c("College or University degree")] <- NA

# ses_household_income_before_tax_avg
ukb_v4$ses_household_income_before_tax_avg.0.0[ukb_v4$ses.accom_type.0.0 %in% c("Sheltered accommodation", "Care home")] <- NA

```

# Save file for imputation

```{r}
saveRDS(ukb_v4, file = "/rds/general/project/hda_25-26/live/TDS/TDS_Group8/imputation/ukb_v5.rds")

```

# One-hot encoding after imputation (mh_activities)

```{r}
ukb_path <- "/rds/general/project/hda_25-26/live/TDS/TDS_Group8/imputation/ukb_train.rds"
ukb_train <- readRDS(ukb_path)

mh_cols_target <- paste0("mh_activities.0.", 0:4)
mh_cols_present <- mh_cols_target[mh_cols_target %in% names(ukb_train)]

mh_mat <- ukb_train[, mh_cols_present, drop = FALSE]
mh_chr <- mh_mat %>%
  mutate(across(everything(), ~ as.character(.)))

mh_mat_chr <- as.matrix(mh_chr)

# Collect unique activity codes (non-missing)
all_codes <- sort(unique(c(mh_mat_chr)))
all_codes <- all_codes[!is.na(all_codes)]

# safe column name helper
safe_code_name <- function(x) {
  x <- as.character(x)
  x <- gsub("[^A-Za-z0-9]+", "_", x)
  x <- gsub("^_+|_+$", "", x)
  x
}

onehot_names <- paste0("mh_activities_", safe_code_name(all_codes))

onehot <- matrix(0L, nrow = nrow(mh_mat_chr), ncol = length(all_codes),
                 dimnames = list(NULL, onehot_names))

for (j in seq_along(all_codes)) {
  code <- all_codes[j]
  onehot[, j] <- as.integer(rowSums(mh_mat_chr == code, na.rm = TRUE) > 0)
}


#bind columns
ukb_train <- cbind(ukb_train,as.data.frame(onehot))

#drop columns
ukb_train<- ukb_train %>%
  select(-all_of(mh_cols_present)) 


#do the same for ukb_test
ukb_test <- readRDS("/rds/general/project/hda_25-26/live/TDS/TDS_Group8/imputation/ukb_test.rds")

mh_cols_target <- paste0("mh_activities.0.", 0:4)
mh_cols_present <- mh_cols_target[mh_cols_target %in% names(ukb_test)]

mh_mat <- ukb_test[, mh_cols_present, drop = FALSE]
mh_chr <- mh_mat %>%
  mutate(across(everything(), ~ as.character(.)))

mh_mat_chr <- as.matrix(mh_chr)

# Collect unique activity codes (non-missing)
all_codes <- sort(unique(c(mh_mat_chr)))
all_codes <- all_codes[!is.na(all_codes)]

# safe column name helper
safe_code_name <- function(x) {
  x <- as.character(x)
  x <- gsub("[^A-Za-z0-9]+", "_", x)
  x <- gsub("^_+|_+$", "", x)
  x
}

onehot_names <- paste0("mh_activities_", safe_code_name(all_codes))

onehot <- matrix(0L, nrow = nrow(mh_mat_chr), ncol = length(all_codes),
                 dimnames = list(NULL, onehot_names))

for (j in seq_along(all_codes)) {
  code <- all_codes[j]
  onehot[, j] <- as.integer(rowSums(mh_mat_chr == code, na.rm = TRUE) > 0)
}


#bind columns
ukb_test <- cbind(ukb_test,as.data.frame(onehot))

#drop columns
ukb_test<- ukb_test %>%
  select(-all_of(mh_cols_present)) 
```

# Deterministic recode (mh_activities)

```{r}
mh_exclusive <- function(df) {
  mh_cols <- grep("^mh_activities_", names(df), value = TRUE)
  
  none_col <- "mh_activities_None_of_the_above"
  pna_col  <- "mh_activities_Prefer_not_to_answer"
  
  other_cols <- setdiff(mh_cols, c(none_col, pna_col))
  
  # if None_of_the_above == 1, set all other mh_activities_* = 0
  idx_none <- df[[none_col]] == 1
  df[idx_none, other_cols] <- 0
  
  # if Prefer_not_to_answer == 1, set all other mh_activities_* = 0
  idx_pna <- df[[pna_col]] == 1
  df[idx_pna, other_cols] <- 0
  
  df
}

ukb_train <- mh_exclusive(ukb_train)
ukb_test <- mh_exclusive(ukb_test)
```

# One-hot encoding & deterministic recode (med_prch)

```{r}
ukb <- ukb_train
# target columns (array 0-5)
med_cols_target <- paste0("med_pain_relief_constipation_heartburn.0.", 0:5)
med_cols_present <- med_cols_target[med_cols_target %in% names(ukb)]

if (length(med_cols_present) == 0) {
  stop("No med_pain_relief_constipation_heartburn.0.* columns found in dataset.")
}

med_mat <- ukb[, med_cols_present, drop = FALSE] %>%
  mutate(across(everything(), as.character))

med_mat_chr <- as.matrix(med_mat)

#Build one-hot indicators
# Collect unique codes (including negative exclusive codes if present)
all_codes <- sort(unique(c(med_mat_chr)))
all_codes <- all_codes[!is.na(all_codes)]

# Safe column name helper
safe_code_name <- function(x) {
  x <- as.character(x)
  x <- str_trim(x)
  x <- str_to_lower(x)
  x <- gsub("[^a-z0-9]+", "_", x)
  x <- gsub("^_+|_+$", "", x)
  x
}

onehot_names <- paste0("med_prch_", safe_code_name(all_codes))

onehot <- matrix(
  0L,
  nrow = nrow(med_mat_chr),
  ncol = length(all_codes),
  dimnames = list(NULL, onehot_names)
)

for (j in seq_along(all_codes)) {
  code <- all_codes[j]
  onehot[, j] <- as.integer(rowSums(med_mat_chr == code, na.rm = TRUE) > 0)
}

#Attach one-hot columns to dataset
onehot_df <- as.data.frame(onehot)
ukb <- bind_cols(ukb, onehot_df)

# rename specified columns
names(ukb)[names(ukb) == "med_prch_ibuprofen_e_g_nurofen"] <- "med_prch_ibuprofen"
names(ukb)[names(ukb) == "med_prch_laxatives_e_g_dulcolax_senokot"] <- "med_prch_laxatives"
names(ukb)[names(ukb) == "med_prch_omeprazole_e_g_zanprol"] <- "med_prch_omeprazole"
names(ukb)[names(ukb) == "med_prch_ranitidine_e_g_zantac"] <- "med_prch_ranitidine"

#Enforce exclusivity
# Columns created by this script
mh_cols <- grep("^med_prch_", names(ukb), value = TRUE)

none_col <- "med_prch_none_of_the_above"
other_cols <- setdiff(mh_cols, none_col)

# if none_of_the_above == 1, all other med_prch_* = 0
idx_none <- ukb[[none_col]] == 1
ukb[idx_none, other_cols] <- 0

#Drop original array columns
ukb <- ukb %>% select(-all_of(med_cols_present))
saveRDS(ukb, "/rds/general/project/hda_25-26/live/TDS/TDS_Group8/imputation/ukb_train.rds")


#Do the same for ukb_test
ukb <- ukb_test
# target columns (array 0-5)
med_cols_target <- paste0("med_pain_relief_constipation_heartburn.0.", 0:5)
med_cols_present <- med_cols_target[med_cols_target %in% names(ukb)]

if (length(med_cols_present) == 0) {
  stop("No med_pain_relief_constipation_heartburn.0.* columns found in dataset.")
}

med_mat <- ukb[, med_cols_present, drop = FALSE] %>%
  mutate(across(everything(), as.character))

med_mat_chr <- as.matrix(med_mat)

#Build one-hot indicators
# Collect unique codes (including negative exclusive codes if present)
all_codes <- sort(unique(c(med_mat_chr)))
all_codes <- all_codes[!is.na(all_codes)]

# Safe column name helper
safe_code_name <- function(x) {
  x <- as.character(x)
  x <- str_trim(x)
  x <- str_to_lower(x)
  x <- gsub("[^a-z0-9]+", "_", x)
  x <- gsub("^_+|_+$", "", x)
  x
}

onehot_names <- paste0("med_prch_", safe_code_name(all_codes))

onehot <- matrix(
  0L,
  nrow = nrow(med_mat_chr),
  ncol = length(all_codes),
  dimnames = list(NULL, onehot_names)
)

for (j in seq_along(all_codes)) {
  code <- all_codes[j]
  onehot[, j] <- as.integer(rowSums(med_mat_chr == code, na.rm = TRUE) > 0)
}

#Attach one-hot columns to dataset
onehot_df <- as.data.frame(onehot)
ukb <- bind_cols(ukb, onehot_df)

# rename specified columns
names(ukb)[names(ukb) == "med_prch_ibuprofen_e_g_nurofen"] <- "med_prch_ibuprofen"
names(ukb)[names(ukb) == "med_prch_laxatives_e_g_dulcolax_senokot"] <- "med_prch_laxatives"
names(ukb)[names(ukb) == "med_prch_omeprazole_e_g_zanprol"] <- "med_prch_omeprazole"
names(ukb)[names(ukb) == "med_prch_ranitidine_e_g_zantac"] <- "med_prch_ranitidine"

#Enforce exclusivity
# Columns created by this script
mh_cols <- grep("^med_prch_", names(ukb), value = TRUE)

none_col <- "med_prch_none_of_the_above"
other_cols <- setdiff(mh_cols, none_col)

# if none_of_the_above == 1, all other med_prch_* = 0
idx_none <- ukb[[none_col]] == 1
ukb[idx_none, other_cols] <- 0

#Drop original array columns
ukb <- ukb %>% select(-all_of(med_cols_present))
saveRDS(ukb, "/rds/general/project/hda_25-26/live/TDS/TDS_Group8/imputation/ukb_test.rds")
```
