---
title: "Aim1"
output: html_document
date: "2026-03-20"
---

# Workflow overview:

0)  Setup
1)  Prepare train and test datasets
2)  LASSO
3)  Stability selection
4)  Make datasets with only baseline variables
5)  Make onehot encoding datasets with selected variables
6)  Make ee_biomarker datasets

### 0) Setup

Load all necessary libraries and datasets

```{r, libraries}
library(dplyr)
library(glmnet)
library(pROC)
library(ggplot2)
library(sharp)
library(data.table)

ukb_train <- readRDS("../pipeline_outputs/post_imputation_output/train_final.rds")
ukb_test  <- readRDS("../pipeline_outputs/post_imputation_output/test_final.rds")

output_dir <- "../pipeline_outputs/aim1_output"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1) Define outcome variable and prediction variables

```{r}
# Outcome must be binary (0/1)
outcome <- "cvd_event"

drop_cols <- c(
  "eid", 
  "date_death.0.0", "date_lost_to_fup.0.0", "date_recr.0.0",
  "assessment_centre_location.0.0"
)

drop_cols <- intersect(drop_cols, names(ukb_train))

baseline_covars <- c("age_recr.0.0", "sex.0.0", "ethnicity.0.0",'bmi','sbp_avg','dbp_avg')
baseline_covars <- intersect(baseline_covars, names(ukb_train))

# keep: external exposome
external_patterns <- c(
  "^env_",
  "^diet_",
  "^smoke_",
  "^phys_",
  "^sed_",
  "^ses_",
  "^mh_",
  "^emp_",
  "^imd$",
  "^sleep_category$",
  "^alcohol_category$",
  "^qualification\\."
)

external_vars <- unique(unlist(lapply(external_patterns, function(p) grep(p, names(ukb_train), value = TRUE))))

preds <- unique(c(baseline_covars, external_vars))
preds <- setdiff(preds, c(outcome, drop_cols))
```

#### 1.1) Prepare training and test datasets

```{r}
train_df <- ukb_train %>% dplyr::select(dplyr::all_of(c(outcome, preds)))
test_df  <- ukb_test  %>% dplyr::select(dplyr::all_of(c(outcome, preds)))

# Ensure outcome is numeric binary (0/1)
train_df[[outcome]] <- as.integer(train_df[[outcome]])
test_df[[outcome]]  <- as.integer(test_df[[outcome]])

y_tr <- train_df[[outcome]]
y_te <- test_df[[outcome]]

# Convert predictors to model matrix (automatic one-hot encoding)
X_tr <- model.matrix(
  as.formula(paste0(outcome, " ~ ", paste(preds, collapse = " + "))),
  data = train_df
)

# Remove intercept column
X_tr <- X_tr[, colnames(X_tr) != "(Intercept)", drop = FALSE]

# Construct test design matrix and align columns with training matrix
X_te <- model.matrix(
  as.formula(paste0(outcome, " ~ ", paste(preds, collapse = " + "))),
  data = test_df
)
X_te <- X_te[, colnames(X_te) != "(Intercept)", drop = FALSE]

# Align test matrix columns to match training matrix
train_cols <- colnames(X_tr)
missing_cols <- setdiff(train_cols, colnames(X_te))
extra_cols   <- setdiff(colnames(X_te), train_cols)

# Add missing columns to test matrix (fill with 0)
if (length(missing_cols) > 0) {
  X_te <- cbind(
    X_te,
    matrix(0, nrow = nrow(X_te), ncol = length(missing_cols),
           dimnames = list(NULL, missing_cols))
  )
}

# Remove extra columns not present in training matrix
if (length(extra_cols) > 0) {
  X_te <- X_te[, setdiff(colnames(X_te), extra_cols), drop = FALSE]
}

# Reorder columns to match training matrix exactly
X_te <- X_te[, train_cols, drop = FALSE]

# Set up for penalty factor
escape_regex <- function(x) {
  gsub("([][{}()+*^$|\\\\?.])", "\\\\\\1", x)
}
penalty_factor <- rep(1, ncol(X_tr))
names(penalty_factor) <- colnames(X_tr)

for (v in baseline_covars) {
  pat <- paste0("^", escape_regex(v))
  penalty_factor[grepl(pat, colnames(X_tr))] <- 0
}

table(penalty_factor)
```

## 2) Fit cross-validated LASSO logistic regression

```{r}
set.seed(123)
cvfit <- cv.glmnet(
  x = X_tr,
  y = y_tr,
  family = "binomial",     # logistic regression
  alpha = 1,               # LASSO penalty
  nfolds = 10,             # 10-fold cross-validation
  type.measure = "auc",    # choose lambda based on AUC
  standardize = TRUE,      # standardize predictors internally
  penalty.factor = penalty_factor
)

lambda_use <- cvfit$lambda.min

# Evaluate performance on test set
p_te <- as.numeric(
  predict(cvfit, newx = X_te, s = lambda_use, type = "response")
)

roc_obj <- pROC::roc(y_te, p_te, quiet = TRUE)
auc_te  <- as.numeric(pROC::auc(roc_obj))

roc_df <- data.frame(
  tpr = roc_obj$sensitivities,
  fpr = 1 - roc_obj$specificities
)

lassoplot <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey") +
  labs(
    title = "ROC Curve - LASSO Logistic Model",
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  annotate("text",
           x = 0.6, y = 0.2,
           label = paste0("AUC = ", round(auc_te, 3)),
           size = 5) +
  theme_minimal(base_size = 14)

ggsave(
  file.path(output_dir, "lasso_ROC_curve.png"),
  plot = lassoplot,
  width = 8,
  height = 6,
  dpi = 300
)

# Extract selected variables (non-zero coefficients)
coef_mat <- as.matrix(coef(cvfit, s = lambda_use))

coef_df <- data.frame(
  term = rownames(coef_mat),
  beta = as.numeric(coef_mat[, 1]),
  row.names = NULL
) %>%
  filter(term != "(Intercept)", beta != 0) %>%
  mutate(abs_beta = abs(beta)) %>%
  arrange(desc(abs_beta))

saveRDS(coef_df, file.path(output_dir, "coef_df.rds"))
```

## 3) Stability selection

```{r}
stab <- VariableSelection(
  xdata = X_tr,
  ydata = y_tr,
  family = "binomial",
  verbose = TRUE,
  penalty.factor = penalty_factor,
  n_cat = 3
)

png(
  filename = "../pipeline_outputs/aim1_output/CalibrationPlot.png",
  width = 2000,
  height = 1600,
  res = 300
)

CalibrationPlot(stab)

dev.off()

saveRDS(stab, file.path(output_dir, "stab.rds"))

# Extract results
sel_prop <- SelectionProportions(stab)

sel_prop_df <- data.frame(
  variable = names(sel_prop),
  selection_proportion = as.numeric(sel_prop)
) %>%
  arrange(desc(selection_proportion))

# Select variables whose selection_proportion > best_pi
best <- Argmax(stab)
best_pi     <- best[1, "pi"]   # 0.52

stable_vars_df <- sel_prop_df %>%
  filter(selection_proportion > best_pi) %>%
  arrange(desc(selection_proportion))

saveRDS(sel_prop_df, "../pipeline_outputs/aim1_output/stable_vars_df.rds")


# Refit final model on selected variables and baseline variables
stable_vars <- stable_vars_df$variable
escape_regex <- function(x) {
  gsub("([][{}()+*^$|\\\\?.])", "\\\\\\1", x)
}

baseline_cols <- unique(unlist(lapply(baseline_covars, function(v) {
  grep(paste0("^", escape_regex(v)), colnames(X_tr), value = TRUE)
})))

final_vars <- unique(c(baseline_cols, stable_vars))

final_vars <- intersect(final_vars, colnames(X_tr))

X_tr_final <- X_tr[, final_vars, drop = FALSE]
X_te_final <- X_te[, final_vars, drop = FALSE]

final_cvfit <- cv.glmnet(
  x = X_tr_final,
  y = y_tr,
  family = 'binomial',
  alpha = 1,
  nfolds = 10,
  type.measure = "auc",
  standardize = TRUE
)

test_prob <- as.numeric(
  predict(final_cvfit, newx = X_te_final, s = "lambda.min", type = "response")
)


# Example AUC
auc_obj <- roc(y_te, test_prob)
auc(auc_obj)

roc_df <- data.frame(
  tpr = auc_obj$sensitivities,
  fpr = 1 - auc_obj$specificities
)

stabsplot <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "blue", size = 1.2) +
  geom_abline(linetype = "dashed", color = "grey") +
  labs(
    title = paste0("ROC Curve (AUC = ", round(auc(auc_obj), 3), ")"),
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  theme_minimal()

ggsave(
  "../pipeline_outputs/aim1_output/stabs_ROC_curve.png",
  plot = stabsplot,
  width = 8,
  height = 6,
  dpi = 300
)
```

## 4) Make datasets with only baseline variables
```{r}
keep_cols <- c("eid", "cvd_event", "age_recr.0.0", "sex.0.0", "ethnicity.0.0", 
               "bmi", "sbp_avg", "dbp_avg")

keep_cols <- intersect(keep_cols, names(ukb_train))

ukb_train_baseline <- ukb_train[, keep_cols, drop = FALSE]
ukb_test_baseline  <- ukb_test[, keep_cols, drop = FALSE]

write.csv(
  ukb_train_baseline,
  "../pipeline_outputs/aim1_output/ukb_train_baseline.csv",
  row.names = FALSE
)

write.csv(
  ukb_test_baseline,
  "../pipeline_outputs/aim1_output/ukb_test_baseline.csv",
  row.names = FALSE
)

```

## 5) Make onehot encoding dataset with selected variables

```{r}
train_final_onehot <- model.matrix(~ . - 1, data = train_df) |> as.data.frame()
test_final_onehot  <- model.matrix(~ . - 1, data = test_df)  |> as.data.frame()

selected_vars <- stable_vars_df$variable
train_baseline_ee_onehot <- train_final_onehot[, selected_vars, drop = FALSE]
test_baseline_ee_onehot <- test_final_onehot[, selected_vars, drop = FALSE]

onehot_extra <- c(
  grep("^bmi", names(test_final_onehot), value = TRUE),
  grep("^sex", names(test_final_onehot), value = TRUE),
  grep("^ethnicity", names(test_final_onehot), value = TRUE)
)

other_covars <- c(
  "eid",
  "cvd_event",
  "age_recr.0.0",
  "sbp_avg",
  "dbp_avg"
)


train_baseline_ee_onehot <- cbind(
  ukb_train[, ..other_covars],
  train_final_onehot[, onehot_extra, drop = FALSE],
  train_baseline_ee_onehot
)

test_baseline_ee_onehot <- cbind(
  ukb_test[, ..other_covars],
  test_final_onehot[, onehot_extra, drop = FALSE],
  test_baseline_ee_onehot
)

train_baseline_ee_onehot <- dplyr::select(train_baseline_ee_onehot, -`sex.0.0Female`)
test_baseline_ee_onehot  <- dplyr::select(test_baseline_ee_onehot, -`sex.0.0Female`)

write.csv(train_baseline_ee_onehot,
          "../pipeline_outputs/aim1_output/train_baseline_ee_onehot.csv",
          row.names = FALSE)

write.csv(test_baseline_ee_onehot,
          "../pipeline_outputs/aim1_output/test_baseline_ee_onehot.csv",
          row.names = FALSE)

```

## 6) Make ee_biomarker datasets
```{r}
bio_vars <- c(
  "blood_haematocrit_percentage.0.0",
  "blood_hb_conc.0.0",
  "blood_high_light_scatter_reticulocyte_count.0.0",
  "blood_high_light_scatter_reticulocyte_percentage.0.0",
  "blood_immature_reticulocyte_fraction.0.0",
  "blood_lymphocyte_count.0.0",
  "blood_lymphocyte_percentage.0.0",
  "blood_mean_platelet_thrombocyte_volume.0.0",
  "blood_mean_reticulocyte_volume.0.0",
  "blood_monocyte_count.0.0",
  "blood_monocyte_percentage.0.0",
  "blood_neutrophill_count.0.0",
  "blood_neutrophill_percentage.0.0",
  "blood_platelet_count.0.0",
  "blood_platelet_crit.0.0",
  "blood_platelet_distribution_width.0.0",
  "blood_rbc_count.0.0",
  "blood_rbc_distribution_width.0.0",
  "blood_reticulocyte_count.0.0",
  "blood_reticulocyte_percentage.0.0",
  "blood_wbc_count.0.0",
  "biochem_apoa.0.0",
  "biochem_apob.0.0",
  "biochem_cholesterol.0.0",
  "biochem_crp.0.0",
  "biochem_glucose.0.0",
  "biochem_hba1c.0.0",
  "biochem_hdl.0.0",
  "biochem_ldl.0.0",
  "biochem_lipoa.0.0",
  "biochem_sodium_urine.0.0",
  "biochem_triglycerides.0.0"
)

train_bio_df <- as.data.frame(ukb_train[, ..bio_vars])
test_bio_df  <- as.data.frame(ukb_test[, ..bio_vars])

train_ee_biomarker_onehot <- cbind(
  as.data.frame(train_baseline_ee_onehot),
  train_bio_df
)

test_ee_biomarker_onehot <- cbind(
  as.data.frame(test_baseline_ee_onehot),
  test_bio_df
)

front_cols <- c("eid", "cvd_event")

train_other_cols <- sort(setdiff(colnames(train_ee_biomarker_onehot), front_cols))
test_other_cols  <- sort(setdiff(colnames(test_ee_biomarker_onehot), front_cols))

train_ee_biomarker_onehot <- train_ee_biomarker_onehot[, c(front_cols, train_other_cols), drop = FALSE]
test_ee_biomarker_onehot  <- test_ee_biomarker_onehot[, c(front_cols, test_other_cols), drop = FALSE]

write.csv(
  train_ee_biomarker_onehot,
  "../pipeline_outputs/aim1_output/train_ee_biomarker_onehot.csv",
  row.names = FALSE
)

write.csv(
  test_ee_biomarker_onehot,
  "../pipeline_outputs/aim1_output/test_ee_biomarker_onehot.csv",
  row.names = FALSE
)

saveRDS(
  train_ee_biomarker_onehot,
  "../pipeline_outputs/aim1_output/train_ee_biomarker_onehot.rds"
)

saveRDS(
  test_ee_biomarker_onehot,
  "../pipeline_outputs/aim1_output/test_ee_biomarker_onehot.rds"
)
```
