---
title: "Aim2_3_Clustering_and_Biomarker_Characterisation"
output: html_document
date: "2026-03-26"
---

## 0) Setup

```{r setup, message=FALSE, warning=FALSE}
rm(list = ls())

library(dplyr)
library(ggplot2)
library(uwot)
library(glmnet)
library(pROC)
library(sharp)

# -------------------------------------------------------------------------
# Define input and output directories
# -------------------------------------------------------------------------
input_dir <- "../pipeline_outputs/aim1_output"
out_dir   <- "../pipeline_outputs/aim2_3_output"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# -------------------------------------------------------------------------
# Load previously prepared datasets
# These datasets are assumed to already contain:
#   1. selected one-hot encoded exposome variables
#   2. biomarker variables
# -------------------------------------------------------------------------
train_dat <- readRDS(file.path(input_dir, "train_ee_biomarker_onehot.rds"))
test_dat  <- readRDS(file.path(input_dir, "test_ee_biomarker_onehot.rds"))

# Check dimensions
dim(train_dat)
dim(test_dat)

# Inspect first few column names
head(names(train_dat))

```

## 1) Define biomarker variables and display labels
```{r}
# -------------------------------------------------------------------------
# Define biomarker variables expected in the saved datasets
# -------------------------------------------------------------------------
bio_vars <- c(
  "blood_haematocrit_percentage.0.0",
  "blood_hb_conc.0.0",
  "blood_high_light_scatter_reticulocyte_count.0.0",
  "blood_high_light_scatter_reticulocyte_percentage.0.0",
  "blood_immature_reticulocyte_fraction.0.0",
  "blood_lymphocyte_count.0.0",
  "blood_lymphocyte_percentage.0.0",
  "blood_mean_platelet_thrombocyte_volume.0.0",
  "blood_mean_reticulocyte_volume.0.0",
  "blood_monocyte_count.0.0",
  "blood_monocyte_percentage.0.0",
  "blood_neutrophill_count.0.0",
  "blood_neutrophill_percentage.0.0",
  "blood_platelet_count.0.0",
  "blood_platelet_crit.0.0",
  "blood_platelet_distribution_width.0.0",
  "blood_rbc_count.0.0",
  "blood_rbc_distribution_width.0.0",
  "blood_reticulocyte_count.0.0",
  "blood_reticulocyte_percentage.0.0",
  "blood_wbc_count.0.0",
  "biochem_apoa.0.0",
  "biochem_apob.0.0",
  "biochem_cholesterol.0.0",
  "biochem_crp.0.0",
  "biochem_glucose.0.0",
  "biochem_hba1c.0.0",
  "biochem_hdl.0.0",
  "biochem_ldl.0.0",
  "biochem_lipoa.0.0",
  "biochem_sodium_urine.0.0",
  "biochem_triglycerides.0.0"
)

# Keep only biomarkers present in both datasets
bio_vars <- Reduce(intersect, list(bio_vars, names(train_dat), names(test_dat)))

# -------------------------------------------------------------------------
# Define biomarker names for tables and plots
# Names = original variable names
# Values = display labels
# -------------------------------------------------------------------------
bio_labels <- c(
  "blood_haematocrit_percentage.0.0" = "Haematocrit",
  "blood_hb_conc.0.0" = "Haemoglobin",
  "blood_high_light_scatter_reticulocyte_count.0.0" = "High light scatter reticulocyte count",
  "blood_high_light_scatter_reticulocyte_percentage.0.0" = "High light scatter reticulocyte percentage",
  "blood_immature_reticulocyte_fraction.0.0" = "Immature reticulocyte fraction",
  "blood_lymphocyte_count.0.0" = "Lymphocyte count",
  "blood_lymphocyte_percentage.0.0" = "Lymphocyte percentage",
  "blood_mean_platelet_thrombocyte_volume.0.0" = "Mean platelet volume",
  "blood_mean_reticulocyte_volume.0.0" = "Mean reticulocyte volume",
  "blood_monocyte_count.0.0" = "Monocyte count",
  "blood_monocyte_percentage.0.0" = "Monocyte percentage",
  "blood_neutrophill_count.0.0" = "Neutrophil count",
  "blood_neutrophill_percentage.0.0" = "Neutrophil percentage",
  "blood_platelet_count.0.0" = "Platelet count",
  "blood_platelet_crit.0.0" = "Plateletcrit",
  "blood_platelet_distribution_width.0.0" = "Platelet distribution width",
  "blood_rbc_count.0.0" = "Red blood cell count",
  "blood_rbc_distribution_width.0.0" = "Red blood cell distribution width",
  "blood_reticulocyte_count.0.0" = "Reticulocyte count",
  "blood_reticulocyte_percentage.0.0" = "Reticulocyte percentage",
  "blood_wbc_count.0.0" = "White blood cell count",
  "biochem_apoa.0.0" = "Apolipoprotein A",
  "biochem_apob.0.0" = "Apolipoprotein B",
  "biochem_cholesterol.0.0" = "Cholesterol",
  "biochem_crp.0.0" = "C-reactive protein",
  "biochem_glucose.0.0" = "Glucose",
  "biochem_hba1c.0.0" = "HbA1c",
  "biochem_hdl.0.0" = "HDL cholesterol",
  "biochem_ldl.0.0" = "LDL cholesterol",
  "biochem_lipoa.0.0" = "Lipoprotein(a)",
  "biochem_sodium_urine.0.0" = "Urinary sodium",
  "biochem_triglycerides.0.0" = "Triglycerides"
)

# Restrict label vector to biomarkers that are actually being used
bio_labels <- bio_labels[bio_vars]

# Helper function:
# returns readable biomarker labels where available,
# otherwise falls back to the original variable name
pretty_bio_names <- function(x) {
  ifelse(x %in% names(bio_labels), bio_labels[x], x)
}
```

###2) Define clustering variables
```{r}
# -------------------------------------------------------------------------
# For clustering, use all columns except:
#   - identifiers / outcomes
#   - biomarker variables
# -------------------------------------------------------------------------
exclude_for_clustering <- c("eid", "cvd_event", bio_vars)

cluster_vars <- setdiff(names(train_dat), exclude_for_clustering)

```

###3) Prepare matrices for clustering
```{r}
# -------------------------------------------------------------------------
# Extract clustering matrices from the selected exposome variables only
# -------------------------------------------------------------------------
X_tr <- as.matrix(train_dat[, cluster_vars, drop = FALSE])
X_te <- as.matrix(test_dat[, cluster_vars, drop = FALSE])

# Ensure all values are numeric
storage.mode(X_tr) <- "numeric"
storage.mode(X_te) <- "numeric"

# Remove zero-variance columns based on training data
keep_cols <- apply(X_tr, 2, sd, na.rm = TRUE) > 0

X_tr <- X_tr[, keep_cols, drop = FALSE]
X_te <- X_te[, colnames(X_tr), drop = FALSE]

# Standardise test data using training-set parameters
X_tr_scaled <- scale(X_tr)

center_vals <- attr(X_tr_scaled, "scaled:center")
scale_vals  <- attr(X_tr_scaled, "scaled:scale")

X_te_scaled <- scale(X_te, center = center_vals, scale = scale_vals)

dim(X_tr_scaled)
dim(X_te_scaled)
```


### 5) choosing number of clusters
```{r}
# -------------------------------------------------------------------------
# Manual silhouette function
# -------------------------------------------------------------------------
manual_silhouette_mean <- function(x, clusters) {
  x <- as.matrix(x)
  clusters <- as.integer(clusters)

  dmat <- as.matrix(dist(x))
  n <- nrow(dmat)
  sil <- numeric(n)

  for (i in seq_len(n)) {
    own_clust <- clusters[i]

    own_idx <- which(clusters == own_clust & seq_len(n) != i)

    if (length(own_idx) == 0) {
      a_i <- 0
    } else {
      a_i <- mean(dmat[i, own_idx])
    }

    other_clusts <- setdiff(unique(clusters), own_clust)

    if (length(other_clusts) == 0) {
      b_i <- 0
    } else {
      mean_dists_other <- sapply(other_clusts, function(cl) {
        mean(dmat[i, clusters == cl])
      })
      b_i <- min(mean_dists_other)
    }

    if (max(a_i, b_i) == 0) {
      sil[i] <- 0
    } else {
      sil[i] <- (b_i - a_i) / max(a_i, b_i)
    }
  }

  mean(sil)
}

# -------------------------------------------------------------------------
# Evaluate candidate values of k on a subsample for efficiency
# -------------------------------------------------------------------------
set.seed(123)

idx <- sample(nrow(X_tr_scaled), min(5000, nrow(X_tr_scaled)))
X_sub <- X_tr_scaled[idx, , drop = FALSE]

wss <- numeric(10)
sil_scores <- rep(NA_real_, 10)

for (k in 2:10) {
  km <- kmeans(X_sub, centers = k, nstart = 20)

  wss[k] <- km$tot.withinss
  sil_scores[k] <- manual_silhouette_mean(X_sub, km$cluster)

  cat(
    "k =", k,
    "| WCSS =", round(wss[k], 2),
    "| Silhouette =", round(sil_scores[k], 4), "\n"
  )
}

k_eval_df <- data.frame(
  k = 2:10,
  WCSS = wss[2:10],
  Silhouette = sil_scores[2:10]
)

print(k_eval_df)

saveRDS(k_eval_df, file.path(out_dir, "k_evaluation.rds"))
write.csv(k_eval_df, file.path(out_dir, "k_evaluation.csv"), row.names = FALSE)


# -------------------------------------------------------------------------
# Save elbow plot
# -------------------------------------------------------------------------
png(
  filename = file.path(out_dir, "elbow_plot.png"),
  width = 1800,
  height = 1400,
  res = 200
)

plot(
  2:10, wss[2:10],
  type = "b",                 # both points and lines
  pch = 19,                  # solid circles
  xlab = "Number of clusters (k)",
  ylab = "Total within-cluster sum of squares",
  main = "Elbow Method"
)

dev.off()


# -------------------------------------------------------------------------
# Save silhouette plot
# -------------------------------------------------------------------------
png(
  filename = file.path(out_dir, "silhouette_plot.png"),
  width = 1800,
  height = 1400,
  res = 200
)

plot(
  2:10, sil_scores[2:10],
  type = "b",
  pch = 19,
  xlab = "Number of clusters (k)",
  ylab = "Average silhouette width",
  main = "Manual Silhouette Method"
)

dev.off()
```

will need to choose number of clusters after run on UKB dataset and then run from here>
### 6) Fit final clustering
```{r}
# -------------------------------------------------------------------------
# Fit final k-means model
# Replace centers = 4 with your chosen value of k after inspection
# -------------------------------------------------------------------------
set.seed(123)

km_final <- kmeans(
  X_tr_scaled,
  centers = 4, #change after run on UK biobank
  nstart = 50,
  iter.max = 100
)

# -------------------------------------------------------------------------
# Function to assign each test observation to the nearest training centroid
# -------------------------------------------------------------------------
assign_cluster <- function(x, centers) {
  apply(x, 1, function(row) {
    which.min(colSums((t(centers) - row)^2))
  })
}

# -------------------------------------------------------------------------
# Assign cluster labels to training and test data
# -------------------------------------------------------------------------
train_cluster <- km_final$cluster
test_cluster  <- assign_cluster(X_te_scaled, km_final$centers)

table(train_cluster)
table(test_cluster)

# -------------------------------------------------------------------------
# Save clustered datasets for downstream analyses
# -------------------------------------------------------------------------
train_clustered <- train_dat
test_clustered  <- test_dat

train_clustered$cluster <- train_cluster
test_clustered$cluster  <- test_cluster

saveRDS(train_clustered, file.path(out_dir, "train_ee_biomarker_clustered.rds"))
saveRDS(test_clustered,  file.path(out_dir, "test_ee_biomarker_clustered.rds"))

write.csv(train_clustered, file.path(out_dir, "train_ee_biomarker_clustered.csv"), row.names = FALSE)
write.csv(test_clustered,  file.path(out_dir, "test_ee_biomarker_clustered.csv"), row.names = FALSE)
```
### 7) Evaluate and visualise clusters
```{r}
# -------------------------------------------------------------------------
# UMAP visualisation of clustering structure (training data)
# -------------------------------------------------------------------------
set.seed(123)

umap_res <- umap(
  X_tr_scaled,
  n_neighbors = 15,
  min_dist = 0.3,
  metric = "euclidean"
)

# -------------------------------------------------------------------------
# Save UMAP object for reproducibility
# -------------------------------------------------------------------------
saveRDS(umap_res, file.path(out_dir, "umap_res.rds"))

# -------------------------------------------------------------------------
# Save UMAP plot to output directory
# -------------------------------------------------------------------------
png(
  filename = file.path(out_dir, "umap_clusters.png"),
  width = 1800,
  height = 1400,
  res = 200
)

plot(
  umap_res,
  col = factor(train_cluster),
  pch = 16,
  xlab = "UMAP1",
  ylab = "UMAP2",
  main = "UMAP of Exposome Variables (coloured by cluster)"
)

legend(
  "topright",
  legend = levels(factor(train_cluster)),
  col = seq_along(levels(factor(train_cluster))),
  pch = 16,
  title = "Cluster"
)

dev.off()


# -------------------------------------------------------------------------
# Heatmap of cluster centroids
# Rows = variables, Columns = clusters
# -------------------------------------------------------------------------
graphics.off()

# Extract cluster centroids from final k-means model
centroid_mat <- km_final$centers

# Transpose: rows = variables, columns = clusters
heat_mat <- t(centroid_mat)

# -------------------------------------------------------------------------
# Select top variables with highest variability across clusters
# -------------------------------------------------------------------------
var_across_clusters <- apply(heat_mat, 1, sd)
top_n <- 20

top_vars <- names(sort(var_across_clusters, decreasing = TRUE))[1:min(top_n, nrow(heat_mat))]

print(top_vars)

# Subset matrix
heat_mat_top <- heat_mat[top_vars, , drop = FALSE]

# Set cluster column names
colnames(heat_mat_top) <- paste0("Cluster ", seq_len(ncol(heat_mat_top)))

# -------------------------------------------------------------------------
# Save heatmap matrix
# -------------------------------------------------------------------------
saveRDS(heat_mat_top, file.path(out_dir, "cluster_centroid_heatmap_matrix.rds"))

write.csv(
  data.frame(variable = rownames(heat_mat_top), heat_mat_top, check.names = FALSE),
  file.path(out_dir, "cluster_centroid_heatmap_matrix.csv"),
  row.names = FALSE
)

# -------------------------------------------------------------------------
# Save heatmap figure
# -------------------------------------------------------------------------
png(
  filename = file.path(out_dir, "cluster_centroid_heatmap.png"),
  width = 2200,
  height = 1800,
  res = 220
)

heatmap(
  heat_mat_top,
  Rowv = TRUE,
  Colv = NA,
  scale = "none",
  labRow = rownames(heat_mat_top),  
  labCol = colnames(heat_mat_top),
  col = colorRampPalette(c("blue", "white", "red"))(100),
  margins = c(10, 25),
  cexRow = 0.8,
  cexCol = 1.2,
  main = paste0("Exposure patterns across ", ncol(heat_mat_top), " clusters")
)

dev.off()
```

###
Aim 3
###

### 1) Biomarker summary by cluster 
```{r}
# -------------------------------------------------------------------------
# Build biomarker data frame using the clustered training dataset
# -------------------------------------------------------------------------
bio_df <- train_clustered[, c(bio_vars, "cluster"), drop = FALSE]

for (v in bio_vars) {
  bio_df[[v]] <- as.numeric(bio_df[[v]])
}

bio_df$cluster <- factor(bio_df$cluster)

bio_mean <- aggregate(
  bio_df[, bio_vars, drop = FALSE],
  by = list(cluster = bio_df$cluster),
  FUN = function(x) mean(x, na.rm = TRUE)
)

print(bio_mean)

saveRDS(bio_mean, file.path(out_dir, "biomarker_cluster_means.rds"))
write.csv(bio_mean, file.path(out_dir, "biomarker_cluster_means.csv"), row.names = FALSE)
```
## 2) ANOVA for biomarker differences across clusters
```{r}
bio_anova <- lapply(bio_vars, function(v) {

  df <- data.frame(
    value = bio_df[[v]],
    cluster = factor(bio_df$cluster)
  )

  model <- aov(value ~ cluster, data = df)
  p_val <- summary(model)[[1]][["Pr(>F)"]][1]

  data.frame(
    biomarker = v,
    p_value = p_val
  )
})

# Combine list into a single data frame
bio_anova <- do.call(rbind, bio_anova)

# Adjust p-values (Benjamini-Hochberg)
bio_anova$p_adj_BH <- p.adjust(bio_anova$p_value, method = "BH")

# Add readable biomarker labels
bio_anova$biomarker_label <- pretty_bio_names(bio_anova$biomarker)

# Order by adjusted p-value
bio_anova <- bio_anova[order(bio_anova$p_adj_BH), ]

# Reorder columns
bio_anova <- bio_anova[, c("biomarker", "biomarker_label", "p_value", "p_adj_BH")]

print(bio_anova)

# Save outputs
saveRDS(bio_anova, file.path(out_dir, "biomarker_anova_results.rds"))
write.csv(bio_anova, file.path(out_dir, "biomarker_anova_results.csv"), row.names = FALSE)
```

### 3) Heatmap of top varying biomarkers across clusters for visualisation
```{r}
# -------------------------------------------------------------------------
# Convert cluster-level biomarker means into a matrix
# Rows = clusters
# Columns = biomarkers
# -------------------------------------------------------------------------
bio_mat <- as.matrix(bio_mean[, -1, drop = FALSE])
rownames(bio_mat) <- paste0("Cluster ", bio_mean$cluster)

# -------------------------------------------------------------------------
# Identify biomarkers with the greatest variation across clusters
# -------------------------------------------------------------------------
bio_var_between <- apply(bio_mat, 2, var, na.rm = TRUE)

top_n <- 15
top_bio <- names(sort(bio_var_between, decreasing = TRUE))[1:min(top_n, length(bio_var_between))]

print(top_bio)

# -------------------------------------------------------------------------
# Scale biomarker means across clusters for visual comparison
# This is for display only, not for inference
# -------------------------------------------------------------------------
bio_mat_scaled <- scale(bio_mat)

# -------------------------------------------------------------------------
# Create row labels using readable biomarker names
# -------------------------------------------------------------------------
row_labs <- pretty_bio_names(top_bio)

# -------------------------------------------------------------------------
# Save the underlying matrix used for the heatmap
# -------------------------------------------------------------------------
bio_heatmap_mat <- t(bio_mat_scaled[, top_bio, drop = FALSE])

saveRDS(
  bio_heatmap_mat,
  file.path(out_dir, "biomarker_heatmap_matrix.rds")
)

write.csv(
  data.frame(
    biomarker = rownames(bio_heatmap_mat),
    bio_heatmap_mat,
    check.names = FALSE
  ),
  file.path(out_dir, "biomarker_heatmap_matrix.csv"),
  row.names = FALSE
)

# -------------------------------------------------------------------------
# Save heatmap figure to output directory
# -------------------------------------------------------------------------
png(
  filename = file.path(out_dir, "biomarker_heatmap_top_varying.png"),
  width = 2200,
  height = 1800,
  res = 220
)

heatmap(
  bio_heatmap_mat,
  Rowv = NA,
  Colv = NA,
  scale = "none",
  labRow = row_labs,
  labCol = colnames(bio_heatmap_mat),
  col = colorRampPalette(c("blue", "white", "red"))(100),
  margins = c(10, 20),
  cexRow = 0.9,
  cexCol = 1.1,
  main = "Biomarker levels across clusters"
)

dev.off()
```
###using lasso and stability for biomarker selection

## 4) Set output directory for one-vs-rest biomarker models
```{r}
# -------------------------------------------------------------------------
# Create a subdirectory for the biomarker one-vs-rest models
# -------------------------------------------------------------------------
model_out_dir <- file.path(out_dir, "cluster_biomarker_one_vs_rest")

if (!dir.exists(model_out_dir)) {
  dir.create(model_out_dir, recursive = TRUE)
}
```

## 5) Prepare biomarker modelling datasets
```{r}
# -------------------------------------------------------------------------
# Build biomarker-only data frames from the clustered datasets
# -------------------------------------------------------------------------
bio_train <- as.data.frame(train_clustered[, c(bio_vars, "cluster"), drop = FALSE])
bio_test  <- as.data.frame(test_clustered[, c(bio_vars, "cluster"), drop = FALSE])

# -------------------------------------------------------------------------
# Ensure cluster labels are factors with consistent levels
# -------------------------------------------------------------------------
bio_train$cluster <- factor(bio_train$cluster)
bio_test$cluster  <- factor(bio_test$cluster, levels = levels(bio_train$cluster))

# -------------------------------------------------------------------------
# Ensure biomarker columns are numeric
# -------------------------------------------------------------------------
for (v in bio_vars) {
  bio_train[[v]] <- as.numeric(bio_train[[v]])
  bio_test[[v]]  <- as.numeric(bio_test[[v]])
}

# -------------------------------------------------------------------------
# Remove biomarkers with zero variance in the training set
# -------------------------------------------------------------------------
keep_bio <- sapply(
  bio_train[, bio_vars, drop = FALSE],
  function(x) sd(x, na.rm = TRUE) > 0
)

bio_vars_use <- bio_vars[keep_bio]

# -------------------------------------------------------------------------
# Restrict to retained biomarkers plus cluster
# -------------------------------------------------------------------------
bio_train <- bio_train[, c(bio_vars_use, "cluster"), drop = FALSE]
bio_test  <- bio_test[, c(bio_vars_use, "cluster"), drop = FALSE]

# -------------------------------------------------------------------------
# Use complete cases only for biomarker model fitting
# -------------------------------------------------------------------------
bio_train_cc <- bio_train[complete.cases(bio_train), , drop = FALSE]
bio_test_cc  <- bio_test[complete.cases(bio_test), , drop = FALSE]

# -------------------------------------------------------------------------
# Convert biomarker data frames to model matrices
# -------------------------------------------------------------------------
X_bio_tr <- as.matrix(bio_train_cc[, bio_vars_use, drop = FALSE])
X_bio_te <- as.matrix(bio_test_cc[, bio_vars_use, drop = FALSE])

cluster_levels <- levels(bio_train_cc$cluster)

dim(X_bio_tr)
dim(X_bio_te)
cluster_levels
```

## 6) Run one-vs-rest biomarker models for each cluster
```{r}
# -------------------------------------------------------------------------
# Initialise storage objects
# -------------------------------------------------------------------------
summary_results <- list()
roc_list <- list()
stable_direction_list <- list()

# -------------------------------------------------------------------------
# Fit one-vs-rest models for each cluster
# -------------------------------------------------------------------------
for (target_cluster in cluster_levels) {

  cat("\n====================================================\n")
  cat("Running biomarker model for Cluster", target_cluster, " vs rest\n")
  cat("====================================================\n")

  # -----------------------------------------------------------------------
  # Define binary outcomes for one-vs-rest classification
  # -----------------------------------------------------------------------
  y_tr <- ifelse(as.character(bio_train_cc$cluster) == target_cluster, 1, 0)
  y_te <- ifelse(as.character(bio_test_cc$cluster)  == target_cluster, 1, 0)

  cat("Training outcome counts:\n")
  print(table(y_tr))

  cat("Test outcome counts:\n")
  print(table(y_te))

  # -----------------------------------------------------------------------
  # Run stability selection
  # -----------------------------------------------------------------------
  set.seed(123)

  stab <- VariableSelection(
    xdata = X_bio_tr,
    ydata = y_tr,
    family = "binomial",
    verbose = TRUE
  )

  saveRDS(
    stab,
    file.path(model_out_dir, paste0("stab_cluster", target_cluster, "_vs_rest.rds"))
  )

  png(
    filename = file.path(model_out_dir, paste0("calibrationplot_cluster", target_cluster, "_vs_rest.png")),
    width = 1800,
    height = 1400,
    res = 200
  )

  CalibrationPlot(stab)

  dev.off()

  # -----------------------------------------------------------------------
  # Extract selection proportions
  # -----------------------------------------------------------------------
  sel_prop <- SelectionProportions(stab)

  sel_prop_df <- data.frame(
    variable = names(sel_prop),
    selection_proportion = as.numeric(sel_prop)
  ) %>%
    arrange(desc(selection_proportion))

  sel_prop_df$variable_label <- pretty_bio_names(sel_prop_df$variable)

  saveRDS(
    sel_prop_df,
    file.path(model_out_dir, paste0("sel_prop_df_cluster", target_cluster, "_vs_rest.rds"))
  )

  write.csv(
    sel_prop_df,
    file.path(model_out_dir, paste0("sel_prop_df_cluster", target_cluster, "_vs_rest.csv")),
    row.names = FALSE
  )

  # -----------------------------------------------------------------------
  # Extract best stability threshold
  # -----------------------------------------------------------------------
  best <- Argmax(stab)
  print(best)

  best_pi <- best[1, "pi"]

  stable_vars_df <- sel_prop_df %>%
    filter(selection_proportion > best_pi) %>%
    arrange(desc(selection_proportion))

  saveRDS(
    stable_vars_df,
    file.path(model_out_dir, paste0("stable_vars_df_cluster", target_cluster, "_vs_rest.rds"))
  )

  write.csv(
    stable_vars_df,
    file.path(model_out_dir, paste0("stable_vars_df_cluster", target_cluster, "_vs_rest.csv")),
    row.names = FALSE
  )

  stable_vars <- intersect(stable_vars_df$variable, colnames(X_bio_tr))

  # -----------------------------------------------------------------------
  # Handle case where too few stable variables are selected
  # -----------------------------------------------------------------------
  if (length(stable_vars) < 2) {

    cat("Fewer than 2 stable biomarkers selected for Cluster", target_cluster, " vs rest\n")

    summary_results[[as.character(target_cluster)]] <- data.frame(
      cluster = target_cluster,
      n_stable_vars = length(stable_vars),
      best_pi = best_pi,
      auc = NA_real_
    )

    next
  }

  # -----------------------------------------------------------------------
  # Refit final LASSO using stable biomarkers only
  # -----------------------------------------------------------------------
  X_tr_final <- X_bio_tr[, stable_vars, drop = FALSE]
  X_te_final <- X_bio_te[, stable_vars, drop = FALSE]

  set.seed(123)

  final_cvfit <- cv.glmnet(
    x = X_tr_final,
    y = y_tr,
    family = "binomial",
    alpha = 1,
    nfolds = 10,
    type.measure = "auc",
    standardize = TRUE
  )

  saveRDS(
    final_cvfit,
    file.path(model_out_dir, paste0("final_cvfit_cluster", target_cluster, "_vs_rest.rds"))
  )

  # -----------------------------------------------------------------------
  # Evaluate model on the test set
  # -----------------------------------------------------------------------
  test_prob <- as.numeric(
    predict(final_cvfit, newx = X_te_final, s = "lambda.min", type = "response")
  )

  roc_obj <- roc(y_te, test_prob, quiet = TRUE)
  auc_val <- as.numeric(auc(roc_obj))

  roc_df <- data.frame(
    tpr = roc_obj$sensitivities,
    fpr = 1 - roc_obj$specificities,
    cluster = paste0("Cluster ", target_cluster, " (AUC = ", round(auc_val, 3), ")")
  )

  roc_list[[as.character(target_cluster)]] <- roc_df

  # -----------------------------------------------------------------------
  # Save individual ROC curve
  # -----------------------------------------------------------------------
  p <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
    geom_line(linewidth = 1.2) +
    geom_abline(linetype = "dashed", color = "grey") +
    labs(
      title = paste0("ROC Curve - Cluster ", target_cluster, " vs rest"),
      x = "False Positive Rate",
      y = "True Positive Rate"
    ) +
    theme_minimal()

  ggsave(
    filename = file.path(model_out_dir, paste0("roc_cluster", target_cluster, "_vs_rest.png")),
    plot = p,
    width = 8,
    height = 6,
    dpi = 300
  )

  # -----------------------------------------------------------------------
  # Extract non-zero coefficients from the final model
  # -----------------------------------------------------------------------
  coef_mat <- as.matrix(coef(final_cvfit, s = "lambda.min"))

  coef_df <- data.frame(
    term = rownames(coef_mat),
    beta = as.numeric(coef_mat[, 1]),
    row.names = NULL
  ) %>%
    filter(term != "(Intercept)", beta != 0) %>%
    mutate(
      term_label = pretty_bio_names(term),
      odds_ratio = exp(beta),
      abs_beta = abs(beta),
      direction = ifelse(
        beta > 0,
        paste0("higher in Cluster ", target_cluster),
        paste0("lower in Cluster ", target_cluster)
      )
    ) %>%
    arrange(desc(abs_beta))

  saveRDS(
    coef_df,
    file.path(model_out_dir, paste0("coef_df_cluster", target_cluster, "_vs_rest.rds"))
  )

  write.csv(
    coef_df,
    file.path(model_out_dir, paste0("coef_df_cluster", target_cluster, "_vs_rest.csv")),
    row.names = FALSE
  )

  # -----------------------------------------------------------------------
  # Create per-cluster summary of stable biomarkers with direction
  # -----------------------------------------------------------------------
stable_direction_df <- stable_vars_df %>%
  dplyr::select(variable, selection_proportion, variable_label) %>%
  dplyr::left_join(
    coef_df %>%
      dplyr::select(
        variable = term,
        beta,
        odds_ratio,
        direction
      ),
    by = "variable"
  ) %>%
  dplyr::filter(!is.na(beta)) %>%
  dplyr::mutate(cluster = target_cluster) %>%
  dplyr::select(
    cluster,
    variable,
    variable_label,
    selection_proportion,
    beta,
    odds_ratio,
    direction
  ) %>%
  dplyr::arrange(desc(selection_proportion))

  stable_direction_list[[as.character(target_cluster)]] <- stable_direction_df

  saveRDS(
    stable_direction_df,
    file.path(model_out_dir, paste0("stable_direction_df_cluster", target_cluster, "_vs_rest.rds"))
  )

  write.csv(
    stable_direction_df,
    file.path(model_out_dir, paste0("stable_direction_df_cluster", target_cluster, "_vs_rest.csv")),
    row.names = FALSE
  )

  # -----------------------------------------------------------------------
  # Store summary results
  # -----------------------------------------------------------------------
  summary_results[[as.character(target_cluster)]] <- data.frame(
    cluster = target_cluster,
    n_stable_vars = length(stable_vars),
    best_pi = best_pi,
    auc = auc_val
  )
}
```

---

## 7) Save summary across all cluster biomarker models
```{r}
summary_df <- bind_rows(summary_results) %>%
  arrange(as.numeric(cluster))

print(summary_df)

saveRDS(
  summary_df,
  file.path(model_out_dir, "summary_all_cluster_models.rds")
)

write.csv(
  summary_df,
  file.path(model_out_dir, "summary_all_cluster_models.csv"),
  row.names = FALSE
)

```

## 8) Save combined summary of stable biomarkers across all clusters
```{r}
if (length(stable_direction_list) > 0) {

  stable_direction_summary_df <- dplyr::bind_rows(stable_direction_list) %>%
    dplyr::arrange(as.numeric(cluster), dplyr::desc(selection_proportion))

  print(stable_direction_summary_df)

  saveRDS(
    stable_direction_summary_df,
    file.path(model_out_dir, "stable_direction_summary_all_clusters.rds")
  )

  write.csv(
    stable_direction_summary_df,
    file.path(model_out_dir, "stable_direction_summary_all_clusters.csv"),
    row.names = FALSE
  )
}
```

## 9) Save overlayed ROC curve across all one-vs-rest models

```{r}
if (length(roc_list) > 0) {

  roc_all <- do.call(rbind, roc_list)

  p_overlay <- ggplot(roc_all, aes(x = fpr, y = tpr, colour = cluster)) +
    geom_line(linewidth = 1.2) +
    geom_abline(linetype = "dashed", colour = "grey") +
    labs(
      title = "Overlayed ROC Curves for Cluster Classification (One-vs-Rest)",
      x = "False Positive Rate",
      y = "True Positive Rate",
      colour = "Model"
    ) +
    theme_minimal()

  ggsave(
    filename = file.path(model_out_dir, "roc_overlay_all_clusters.png"),
    plot = p_overlay,
    width = 8.5,
    height = 6.5,
    dpi = 300
  )
}
```


## 10) Save reduced train and test datasets for downstream analyses
```{r}
final_data_dir <- file.path(out_dir, "final_datasets")
dir.create(final_data_dir, recursive = TRUE, showWarnings = FALSE)

# -------------------------------------------------------------------------
# Extract the union of biomarkers stably selected across all clusters
# -------------------------------------------------------------------------
stable_biomarkers_all <- unique(stable_direction_summary_df$variable)

# Keep only biomarkers that are actually present in the clustered datasets
stable_biomarkers_all <- intersect(stable_biomarkers_all, names(train_clustered))

# -------------------------------------------------------------------------
# Define the selected exposome variables to retain
# These are the variables used for clustering after zero-variance filtering
# -------------------------------------------------------------------------
selected_ee_vars <- colnames(X_tr)

# Keep only those that are present in both clustered datasets
selected_ee_vars <- Reduce(
  intersect,
  list(selected_ee_vars, names(train_clustered), names(test_clustered))
)

# -------------------------------------------------------------------------
# Define core columns to retain
# -------------------------------------------------------------------------
core_cols <- c("eid", "cvd_event", "cluster")
core_cols <- Reduce(
  intersect,
  list(core_cols, names(train_clustered), names(test_clustered))
)

# -------------------------------------------------------------------------
# Combine all columns to retain in the final reduced datasets
# -------------------------------------------------------------------------
final_reduced_cols <- unique(c(
  core_cols,
  selected_ee_vars,
  stable_biomarkers_all
))

# Keep only columns present in both datasets
final_reduced_cols <- Reduce(
  intersect,
  list(final_reduced_cols, names(train_clustered), names(test_clustered))
)

# -------------------------------------------------------------------------
# Subset train and test datasets
# -------------------------------------------------------------------------
train_ee_bio_stable_allclusters <- train_clustered[, final_reduced_cols, drop = FALSE]
test_ee_bio_stable_allclusters  <- test_clustered[, final_reduced_cols, drop = FALSE]

# -------------------------------------------------------------------------
# Reorder columns so key identifiers come first
# -------------------------------------------------------------------------
front_cols <- intersect(c("eid", "cvd_event", "cluster"), final_reduced_cols)
other_cols <- setdiff(final_reduced_cols, front_cols)

train_ee_bio_stable_allclusters <- train_ee_bio_stable_allclusters[, c(front_cols, other_cols), drop = FALSE]
test_ee_bio_stable_allclusters  <- test_ee_bio_stable_allclusters[, c(front_cols, other_cols), drop = FALSE]

# -------------------------------------------------------------------------
# Inspect dimensions
# -------------------------------------------------------------------------
dim(train_ee_bio_stable_allclusters)
dim(test_ee_bio_stable_allclusters)

# -------------------------------------------------------------------------
# Save reusable downstream datasets
# -------------------------------------------------------------------------
saveRDS(
  train_ee_bio_stable_allclusters,
  file.path(final_data_dir, "train_ee_bio_stable_allclusters.rds")
)

saveRDS(
  test_ee_bio_stable_allclusters,
  file.path(final_data_dir, "test_ee_bio_stable_allclusters.rds")
)

write.csv(
  train_ee_bio_stable_allclusters,
  file.path(final_data_dir, "train_ee_bio_stable_allclusters.csv"),
  row.names = FALSE
)

write.csv(
  test_ee_bio_stable_allclusters,
  file.path(final_data_dir, "test_ee_bio_stable_allclusters.csv"),
  row.names = FALSE
)

# -------------------------------------------------------------------------
# Save variable lists for reproducibility
# -------------------------------------------------------------------------
saveRDS(
  stable_biomarkers_all,
  file.path(final_data_dir, "stable_biomarkers_allclusters.rds")
)

write.csv(
  data.frame(variable = stable_biomarkers_all),
  file.path(final_data_dir, "stable_biomarkers_allclusters.csv"),
  row.names = FALSE
)

saveRDS(
  selected_ee_vars,
  file.path(final_data_dir, "selected_ee_vars_used_for_clustering.rds")
)

write.csv(
  data.frame(variable = selected_ee_vars),
  file.path(final_data_dir, "selected_ee_vars_used_for_clustering.csv"),
  row.names = FALSE
)
```

