---
title: "Preprocessing"
author: "Group_8"
date: "2026-03-01"
output:
  html_document: default
  pdf_document: default
---

# Workflow overview:

0)  Setup
1)  Distribution checks
2)  Exclusion of individuals with CVD before baseline
3)  Conversion of "Do not know" (IDK) and "Prefer not to answer" (PNA) to NA
4)  Reclassify ethnicity
5)  Removal of outliers
6)  Instance selection
7)  Merging/dropping of variables
8)  IMD tertiles
9)  Alcohol & BMI cats
10) Missingness checks
11) Vars exclusion
12) Table1
13) Imputation

## 0) Setup

Load all necessary libraries

```{r, libraries}
library(dplyr)
library(lubridate)
library(stringr)
```

Load data

```{r, data}
ukb <- readRDS("../extraction_and_recoding/outputs/ukb_recoded.rds")
cvd_event <- readRDS("cvd_events.rds")
```

Generate output dir
```{r}
output_dir <- "../pipeline_outputs/pre_output"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```


Addition of CVD_event, CVD_date and EID cols to data

```{r, prelim}

# add eid column
ukb$eid <- rownames(ukb)
rownames(ukb) <- NULL
ukb$eid <- as.character(ukb$eid)

cvd_event$eid <- as.character(cvd_event$eid)

# cvd event flag
cvd_ids <- unique(cvd_event$eid)
ukb$cvd_event <- as.integer(ukb$eid %in% cvd_ids)

# first cvd date 
first_cvd <- aggregate(date ~ eid, cvd_event, min)

ukb$cvd_date <- first_cvd$date[match(ukb$eid, first_cvd$eid)]

# reorder columns
first_cols <- c("eid", "cvd_event", "cvd_date")
other_cols <- setdiff(names(ukb), first_cols)
other_cols_sorted <- sort(other_cols)
new_order <- c(first_cols, other_cols_sorted)
ukb <- ukb[, new_order]

# check binary
table(ukb$cvd_event, useNA = "ifany")
```

## 1) Distribution checks

Distribution checks are ran on the dataset before any changes. /rds/general/project/hda_25-26/live/TDS/TDS_Group8/preprocessing/distribution_check/dis_full.html - contains all the distribution checks of this dataset at baseline.

## 2) Exclusion

Exclude individuals with CVD_event before recruitment and within 1 year washout period


```{r, exclusion}

ukb_v2 <- ukb %>%
  mutate(
    date_recr.0.0 = as.Date(date_recr.0.0),
    cvd_date      = as.Date(cvd_date),
    
    cvd_case = case_when(
      !is.na(cvd_date) &
        cvd_date <= date_recr.0.0 + years(1) ~ "exclude_washout",
      
      !is.na(cvd_date) &
        cvd_date > date_recr.0.0 + years(1) ~ "incident",
      
      TRUE ~ "no_event"
    )
  )

table(ukb_v2$cvd_case)

# Exclude washout cases
ukb_v2 <- ukb_v2 %>%
  filter(cvd_case != "exclude_washout") %>%
  dplyr::select(-cvd_case)

table(ukb_v2$cvd_event)

# drop cvd_date after washout step
ukb_v2 <- ukb_v2 %>%
  dplyr::select(-cvd_date)
```

## 3) PNA & IDK to NA conversion

Convert all O First check the number of PNA and IDKs throughout the entire dataset.

```{r, check}
N <- nrow(ukb_v2)

# Helper function to compute rates
compute_rate <- function(df, target_string) {
  # Rate among non-NA
  rate_nonNA <- sapply(df, function(x) {
    if (is.character(x) || is.factor(x)) {
      mean(as.character(x) == target_string, na.rm = TRUE)
    } else NA_real_
  })
  
  df_nonNA <- data.frame(
    column = names(rate_nonNA),
    rate_nonNA = as.numeric(rate_nonNA),
    row.names = NULL
  )
  
  # Rate over total rows
  rate_total <- sapply(df, function(x) {
    if (is.character(x) || is.factor(x)) {
      sum(as.character(x) == target_string, na.rm = TRUE) / N
    } else NA_real_
  })
  
  df_total <- data.frame(
    column = names(rate_total),
    rate_total = as.numeric(rate_total),
    row.names = NULL
  )
  
  merged <- merge(
    df_total,
    df_nonNA,
    by = "column",
    all = TRUE,
    suffixes = c("_overall", "_nonNA")
  )
  
  names(merged)[names(merged) == "rate_total"] <- paste0(target_string, "_rate_overallN")
  names(merged)[names(merged) == "rate_nonNA"] <- paste0(target_string, "_rate_among_nonmissing")
  
  merged
}

# Compute PNA rates
pna_table <- compute_rate(ukb_v2, "Prefer not to answer")

# Compute IDK rates
idk_table <- compute_rate(ukb_v2, "I don't know")

# Merge both PNA and IDK tables
merged_table <- merge(pna_table, idk_table, by = "column", all = TRUE)

# Order by highest overall PNA rate (optional)
merged_table <- merged_table[order(-merged_table$`Prefer not to answer_rate_overallN`), ]

# View result
head(merged_table)

```

#### Convert IDK and PNA to NAs

```{r, NA}
#change all IDK to NAs
ukb_v2[] <- lapply(ukb_v2, function(x) {
  if (is.character(x) || is.factor(x)) {
    x[x == "Do not know"] <- NA
  }
  x
})

#check
sum(sapply(ukb_v2, function(x) {
  if (is.factor(x)) x <- as.character(x)
  if (is.character(x)) sum(x == "Do not know", na.rm = TRUE) else 0
}))


# Convert all "Prefer not to answer" to NA
ukb_v2[] <- lapply(ukb_v2, function(x) {
  if (is.character(x) || is.factor(x)) {
    x[x == "Prefer not to answer"] <- NA
  }
  x
})

# Quick check: should be 0 now
sum(sapply(ukb_v2, function(x) {
  if (is.factor(x)) x <- as.character(x)
  if (is.character(x)) sum(x == "Prefer not to answer", na.rm = TRUE) else 0
}))
```

#### Convert all chara vars to numeric

```{r, char-num}

#change pna to na in continuous variables
char_cols <- names(ukb_v2)[sapply(ukb_v2, is.character)]

#change character to numeric
ukb_v2[char_cols] <- lapply(ukb_v2[char_cols], function(x) {
  x[x == "Prefer not to answer"] <- NA
  suppressWarnings(as.numeric(x))
})

#final check
types_df <- data.frame(
  variable = names(ukb_v2),
  class = sapply(ukb_v2, function(x) paste(class(x), collapse = "/")),
  stringsAsFactors = FALSE
)

```

## 4) Reclassify ethnicity

```{r}
eth <- as.character(ukb_v2$ethnicity.0.0)
eth <- trimws(eth)


white_set <- c(
  "White",
  "British", "Irish", "Any other white background"
)

asian_set <- c(
  "Asian or Asian British",
  "Indian", "Pakistani", "Bangladeshi",
  "Any other Asian background",
  "Chinese"
)

mixed_set <- c(
  "Mixed",
  "White and Black Caribbean", "White and Black African", "White and Asian",
  "Any other mixed background"
)

black_set <- c("Black or Black British", "Caribbean", "African", "Any other Black background")

other_set <- c('Other ethnic group')


ukb_v2$ethnicity.0.0 <- dplyr::case_when(
  is.na(eth) ~ NA_character_,
  eth %in% white_set ~ "White",
  eth %in% asian_set ~ "Asian",
  eth %in% mixed_set ~ "Mixed",
  eth %in% black_set ~ "Black",
  eth %in% other_set ~ "Other")

ukb_v2$ethnicity.0.0  <- as.factor(ukb_v2$ethnicity.0.0)
```

## 5) Outlier removal

*For ukbiobank edit this so all values "\< 0" becomes NA*

```{r, outliers}

ukb_v2 <- ukb_v2 %>%
  mutate(across(
    where(is.numeric),
    ~ ifelse(. < -1e5, NA, .)
  ))

```

## 6) Instance selection

Select the latest instance for all vars

*For ukbiobank need to check how instances are ordered compared with synthetic, i.e., is each col one category?*

```{r, instance}
# only have one date of death col (both cols are the exact same)

ukb_v2$date_death.0.0.1 <- NULL 

# keep latest ses_current_employment_status

ukb_v2 <- ukb_v2[, !grepl("ses_current_employment_status\\.0\\.[0-5]$", names(ukb_v2))]

# keep the latest ses_commute_transport_type

ukb_v2 <- ukb_v2[, !grepl("ses_commute_transport_type\\.0\\.[0-2]$", names(ukb_v2))]

# take the latest qualification col

ukb_v2 <- ukb_v2[, !grepl("qualification\\.0\\.[0-4]$", names(ukb_v2))]

```

## 7) Merging cols

```{r, merge}

# for reproducibility take automated sbp and dbp cols

# remove manual bp cols

ukb_v2 <- ukb_v2 %>% 
  dplyr::select(-sbp_manual.0.0, -sbp_manual.0.1, -dbp_manual.0.0, -dbp_manual.0.1)

# average sbp cols
ukb_v2 <- ukb_v2 %>%
  dplyr::mutate(
    dplyr::across(c(sbp_automated.0.0, sbp_automated.0.1), ~ {
      x <- as.numeric(as.character(.))
      x[x == 8242] <- NA
      x
    }),
    sbp_avg = rowMeans(dplyr::across(c(sbp_automated.0.0, sbp_automated.0.1)), na.rm = TRUE)
  ) %>%
  dplyr::select(-sbp_automated.0.0, -sbp_automated.0.1)

# average dbp cols
ukb_v2 <- ukb_v2 %>% 
  dplyr::mutate(
    dbp_avg = rowMeans(dplyr::across(c(dbp_automated.0.0, dbp_automated.0.1)), na.rm = TRUE)
  ) %>%
  dplyr::select(-dbp_automated.0.0, -dbp_automated.0.1)

# drop data
ukb_v2 <- ukb_v2 %>%
  dplyr::select(
    -sleep_snore.1.0, -sleep_snore.2.0, -sleep_snore.3.0,
    -sleep_chronotype.1.0, -sleep_chronotype.2.0, -sleep_chronotype.3.0
  )
```


## 8) IMD tertiles

```{r, imd}

process_imd <- function(df) {
  
  # Map assessment centre to country
  
  england_centres <- c(
    "Stockport (pilot)", "Manchester", "Oxford", "Stoke", "Reading",
    "Bury", "Newcastle", "Leeds", "Bristol", "Barts", "Nottingham",
    "Sheffield", "Liverpool", "Middlesborough", "Hounslow",
    "Croydon", "Birmingham", "Cheadle (revisit)", "Cheadle (imaging)",
    "Reading (imaging)", "Newcastle (imaging)", "Bristol (imaging)"
  )
  
  scotland_centres <- c("Glasgow", "Edinburgh")
  
  wales_centres <- c("Cardiff", "Swansea", "Wrexham")
  
  df <- df %>%
    mutate(
      uk_country = case_when(
        assessment_centre_location.0.0 %in% england_centres  ~ "England",
        assessment_centre_location.0.0 %in% scotland_centres ~ "Scotland",
        assessment_centre_location.0.0 %in% wales_centres    ~ "Wales",
        TRUE ~ NA_character_
      )
    )
  
  # Country specific IMD vars
  
  df <- df %>%
    mutate(
      imd_raw = case_when(
        uk_country == "England"  ~ ses_imd_england.0.0,
        uk_country == "Scotland" ~ ses_imd_scotland.0.0,
        uk_country == "Wales"    ~ ses_imd_wales.0.0,
        TRUE ~ NA_real_
      )
    )
  
  # Country specific IMD tertiles
  
  df <- df %>%
    group_by(uk_country) %>%
    mutate(
      imd_tertile = ntile(imd_raw, 3)
    ) %>%
    ungroup()
  
  # Merge IMD var
  
  df <- df %>%
    mutate(
      imd = imd_tertile
    )
  
  df <- df%>%
    dplyr::select(-uk_country, -imd_raw, -imd_tertile, -ses_imd_england.0.0, -ses_imd_scotland.0.0, -ses_imd_wales.0.0, -assessment_centre_location.0.0)
  
  return(df)
}

# remove assessment centre location (no longer needed)

ukb_v2 <- process_imd(ukb_v2)

```

## 9) Alcohol and BMI cats

```{r, cats}
process_alcohol_bmi <- function(df) {
  
  df %>%
    dplyr::mutate(
      
      #aclohol categories
      alcohol_category = case_when(
        
        # Never drinkers
        alc_freq.0.0 == "Never" ~ "Non-drinker",
        alc_status.0.0 == "Never" ~ "Non-drinker",
        
        # Former drinkers
        alc_status.0.0 == "Previous" ~ "Former",
        
        # Current drinkers by frequency
        alc_status.0.0 == "Current" &
          alc_freq.0.0 %in% c("Special occasions only",
                              "One to three times a month") ~ "Social drinker",
        
        alc_status.0.0 == "Current" &
          alc_freq.0.0 == "Once or twice a week" ~ "Moderate drinker",
        
        alc_status.0.0 == "Current" &
          alc_freq.0.0 %in% c("Three or four times a week",
                              "Daily or almost daily") ~ "Daily drinker",
        
        TRUE ~ NA_character_
      ),
      
      # bmi cats
      bmi = cut(
        bmi.0.0,
        breaks = c(-Inf, 18.5, 25, 30, 40, Inf),
        labels = c(
          "Underweight (<18.5)",
          "Normal weight (18.5–24.9)",
          "Overweight (25–29.9)",
          "Obese (30–39.9)",
          "Severely obese (≥40)"
        ),
        right = FALSE
      )
    ) %>%
    
# drop original cols
  dplyr::select(-bmi.0.0)
}  

ukb_v2 <- process_alcohol_bmi(ukb_v2)
```


## 10) Missingness checks

```{r, missingness}

# Check column-wise missingness

column_missing_rate <- sapply(ukb_v2, function(x) mean(is.na(x)))

column_missing_summary <- data.frame(
  var = names(ukb_v2),
  column_missing_rate = column_missing_rate
)

# pull high missing vars 

high_missing <- column_missing_summary %>%
  filter(column_missing_rate > 0.5) %>%
  mutate(
    percent_missing = round(column_missing_rate * 100, 1)
  ) %>%
  arrange(desc(percent_missing))

high_missing
# Check row-wise missingness

exclude_vars <- intersect(
  c("cvd_date", "date_death.0.0", "date_lost_to_fup.0.0"),
  names(ukb_v2)
)

# Create subset excluding those columns
ukb_subset <- ukb_v2[, !(names(ukb_v2) %in% exclude_vars)]

# Row-wise missing rate
row_missing_rate <- apply(ukb_subset, 1, function(x) mean(is.na(x)))

# Optional: number of missing per row
row_missing_n <- apply(ukb_subset, 1, function(x) sum(is.na(x)))

# Combine into summary dataframe
row_missing_summary <- data.frame(
  missing_n = row_missing_n,
  missing_rate = row_missing_rate
)

# visualising missingness
# column summary
summary(column_missing_rate)
hist(column_missing_rate, breaks = 50)

# row summary
summary(row_missing_rate)
hist(row_missing_rate, breaks = 50)

```

## 11) Excluding vars

Vars have been excluded based on their level of missingness in the UK Biobank. Vars with missingness >50% are excluded.

```{r, exc}

ukb_v2 <- ukb_v2%>%
  dplyr::select( -starts_with("smoke_age"), 
         -starts_with("smoke_tobacco_type_"), -diet_carb.0.0, -diet_energy.0.0, -diet_fat.0.0,
         -diet_food_weight.0.0, -diet_polyunsat_fat.0.0, -diet_protein.0.0, -diet_sat_fat.0.0,
         -diet_total_sugar.0.0, -sleep_wake_early.0.0, -sleep_too_much.0.0, 
         -sleep_falling_asleep_trouble.0.0, -sleep_change.0.0, -sleep_daytime_nap.0.0, 
         -smoke_people_in_household.0.0, -smoke_tobacco_exposure_home.0.0,  
         -smoke_tobacco_exposure_outside.0.0)

```


## 12) Table 1
```{r}
# ---- helper functions ----
fmt_mean_sd <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  paste0(round(mean(x), digits), " (", round(sd(x), digits), ")")
}

fmt_n <- function(x) {
  sum(!is.na(x))
}

cat_table <- function(x, varname, digits = 1, include_na = FALSE) {
  if (!include_na) {
    tab <- table(x, useNA = "no")
    denom <- sum(tab)
  } else {
    tab <- table(x, useNA = "ifany")
    denom <- sum(tab)
  }
  prop <- (tab / denom) * 100
  
  data.frame(
    Variable = varname,
    Level    = names(tab),
    Summary  = paste0(as.integer(tab), " (", round(as.numeric(prop), digits), "%)"),
    stringsAsFactors = FALSE
  )
}

# ---- specify variables ----
cont_vars <- c("age_recr.0.0", "bmi.0.0", "sbp_avg")

cat_vars <- c("sex.0.0",
              "ethnicity.0.0",
              "cvd_event",
              "smoke_status.0.0",
              "alc_status.0.0",
              "ses_current_employment_status.0.6",
              "imd",
              "qualification.0.5")

# ---- build table 1 ----
table1_list <- list()

# add overall N
table1_list[["N_total"]] <- data.frame(
  Variable = "Participants",
  Level = "",
  Summary = as.character(nrow(ukb_v2)),
  stringsAsFactors = FALSE
)

# continuous vars: Mean (SD)
for (v in cont_vars) {
  x <- ukb_v2[[v]]
  table1_list[[v]] <- data.frame(
    Variable = v,
    Level = "",
    Summary = paste0(fmt_mean_sd(x), "; n=", fmt_n(x)),
    stringsAsFactors = FALSE
  )
}

# categorical vars: n (%)
for (v in cat_vars) {
  x <- ukb_v2[[v]]
  table1_list[[v]] <- cat_table(x, varname = v, digits = 1, include_na = FALSE)
}

table1_df <- do.call(rbind, table1_list)

# view in console
print(table1_df, row.names = FALSE)

# export
write.csv(table1_df, file.path(output_dir, "Table1_baseR.csv"), row.names = FALSE)
```

## 12) Imputation

For imputation, run the corresponding imputation script in `pipeline_scripts/`.

This step splits the dataset into training and test sets, fits the imputation model on the training data, and applies the trained model to the test data.

Imputed datasets are saved to the designated output directory.

##### Save

```{r, save}
saveRDS(ukb_v2, file = file.path(output_dir, "ukb_raw.rds"))

```
