---
title: "Preprocessing"
author: "Group_8"
date: "2026-03-01"
output:
  html_document: default
  pdf_document: default
---

# Workflow overview:

0)  Setup
1)  Distribution checks
2)  Exclusion of individuals with CVD before baseline
3)  Conversion of "Do not know" (IDK) and "Prefer not to answer" (PNA) to NA
4)  Reclassify ethnicity
5)  Removal of outliers
6)  Instance selection
7)  Merging/dropping of variables
8)  IMD tertiles
9)  Alcohol cats
10) Missingness checks
11) Vars exclusion
12) Table1
13) Imputation

## 0) Setup

Load all necessary libraries

```{r, libraries}
library(dplyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(stringr)
library(tidyr)
library(tableone)
```

Load data

```{r, data}
ukb <- readRDS("../extraction_and_recoding/outputs/ukb_recoded.rds") 
cvd_event <- readRDS("cvd_events.rds") # update with ukbiobank cvd_event.rds 
```

Generate output dir
```{r}
output_dir <- "../pipeline_outputs/pre_output"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```


Addition of CVD_event, CVD_date and EID cols to data

```{r, prelim}

# add eid column
ukb$eid <- rownames(ukb)
rownames(ukb) <- NULL
ukb$eid <- as.character(ukb$eid)

cvd_event$eid <- as.character(cvd_event$eid)

# cvd event flag
cvd_ids <- unique(cvd_event$eid)
ukb$cvd_event <- as.integer(ukb$eid %in% cvd_ids)

# first cvd date 
first_cvd <- aggregate(date ~ eid, cvd_event, min)

ukb$cvd_date <- first_cvd$date[match(ukb$eid, first_cvd$eid)]

# reorder columns
first_cols <- c("eid", "cvd_event", "cvd_date")
other_cols <- setdiff(names(ukb), first_cols)
other_cols_sorted <- sort(other_cols)
new_order <- c(first_cols, other_cols_sorted)
ukb <- ukb[, new_order]

# check binary
table(ukb$cvd_event, useNA = "ifany")
```

## 1) Distribution checks

Creates a pdf with all variable distributions checked

```{r, dist}

plot_distributions <- function(df,
                                   id_col = "eid",
                                   output_dir = "../pipeline_outputs/pre_output",
                                   output_file = "variable_distributions.pdf",
                                   max_cat_levels = 30,
                                   max_plot_levels = 50,
                                   hist_bins = 30) {
  
  dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
  
  vars <- setdiff(colnames(df), id_col)
  pdf_path <- file.path(output_dir, output_file)
  
  pdf(pdf_path, width = 8, height = 5)
  on.exit(dev.off(), add = TRUE)
  
  for (v in vars) {
    message("Plotting: ", v)
    
    x <- df[[v]]
    
    if (all(is.na(x))) {
      message("Skipping ", v, " because all values are NA")
      next
    }
    
    is_cat <- is.factor(x) || is.character(x) || is.logical(x)
    n_unique <- length(unique(na.omit(x)))
    
    if (!is_cat && is.numeric(x) && n_unique <= max_cat_levels) {
      is_cat <- TRUE
    }
    
    if (is_cat) {
      if (n_unique > max_plot_levels) {
        message("Skipping ", v, " because it has ", n_unique, " levels")
        next
      }
      
      plot_df <- data.frame(plot_var = as.factor(x))
      
      p <- ggplot(plot_df, aes(x = plot_var)) +
        geom_bar(fill = "grey40") +
        coord_flip() +
        labs(
          title = paste0("Plotted from: ", v, " (categorical)"),
          x = NULL,
          y = "Count"
        ) +
        theme_grey()
      
    } else {
      plot_df <- data.frame(plot_var = x)
      plot_df <- plot_df[!is.na(plot_df$plot_var), , drop = FALSE]
      
      p <- ggplot(plot_df, aes(x = plot_var)) +
        geom_histogram(
          bins = hist_bins,
          fill = "grey40",
          color = "grey40"
        ) +
        labs(
          title = paste0("Plotted from: ", v, " (numeric)"),
          x = NULL,
          y = "Count"
        ) +
        theme_grey()
    }
    
    print(p)
  }
  
  message("All plots saved in: ", normalizePath(pdf_path))
}


plot_distributions(ukb, id_col = "eid")

```

## 2) Exclusion

Exclude individuals with CVD_event before recruitment and within 1 year washout period


```{r, exclusion}

ukb_v2 <- ukb %>%
  mutate(
    date_recr.0.0 = as.Date(date_recr.0.0),
    cvd_date      = as.Date(cvd_date),
    
    cvd_case = case_when(
      !is.na(cvd_date) &
        cvd_date <= date_recr.0.0 + years(1) ~ "exclude_washout",
      
      !is.na(cvd_date) &
        cvd_date > date_recr.0.0 + years(1) ~ "incident",
      
      TRUE ~ "no_event"
    )
  )

table(ukb_v2$cvd_case)

# Exclude washout cases
ukb_v2 <- ukb_v2 %>%
  filter(cvd_case != "exclude_washout") %>%
  dplyr::select(-cvd_case)

table(ukb_v2$cvd_event)

# drop cvd_date & date_recr after washout step
ukb_v2 <- ukb_v2 %>%
  dplyr::select(-cvd_date, -date_recr.0.0)
```

## 3) PNA & IDK to NA conversion

Convert all O First check the number of PNA and IDKs throughout the entire dataset.

```{r, check}
N <- nrow(ukb_v2)

# Helper function to compute rates
compute_rate <- function(df, target_string) {
  # Rate among non-NA
  rate_nonNA <- sapply(df, function(x) {
    if (is.character(x) || is.factor(x)) {
      mean(as.character(x) == target_string, na.rm = TRUE)
    } else NA_real_
  })
  
  df_nonNA <- data.frame(
    column = names(rate_nonNA),
    rate_nonNA = as.numeric(rate_nonNA),
    row.names = NULL
  )
  
  # Rate over total rows
  rate_total <- sapply(df, function(x) {
    if (is.character(x) || is.factor(x)) {
      sum(as.character(x) == target_string, na.rm = TRUE) / N
    } else NA_real_
  })
  
  df_total <- data.frame(
    column = names(rate_total),
    rate_total = as.numeric(rate_total),
    row.names = NULL
  )
  
  merged <- merge(
    df_total,
    df_nonNA,
    by = "column",
    all = TRUE,
    suffixes = c("_overall", "_nonNA")
  )
  
  names(merged)[names(merged) == "rate_total"] <- paste0(target_string, "_rate_overallN")
  names(merged)[names(merged) == "rate_nonNA"] <- paste0(target_string, "_rate_among_nonmissing")
  
  merged
}

# Compute PNA rates
pna_table <- compute_rate(ukb_v2, "Prefer not to answer")

# Compute IDK rates
idk_table <- compute_rate(ukb_v2, "I don't know")

# Merge both PNA and IDK tables
merged_table <- merge(pna_table, idk_table, by = "column", all = TRUE)

# Order by highest overall PNA rate (optional)
merged_table <- merged_table[order(-merged_table$`Prefer not to answer_rate_overallN`), ]

# View result
head(merged_table)

```

#### Convert IDK and PNA to NAs

```{r, NA}
#change all IDK to NAs
ukb_v2[] <- lapply(ukb_v2, function(x) {
  if (is.character(x) || is.factor(x)) {
    x[x == "Do not know"] <- NA
  }
  x
})

#check
sum(sapply(ukb_v2, function(x) {
  if (is.factor(x)) x <- as.character(x)
  if (is.character(x)) sum(x == "Do not know", na.rm = TRUE) else 0
}))


# Convert all "Prefer not to answer" to NA
ukb_v2[] <- lapply(ukb_v2, function(x) {
  if (is.character(x) || is.factor(x)) {
    x[x == "Prefer not to answer"] <- NA
  }
  x
})

# Quick check: should be 0 now
sum(sapply(ukb_v2, function(x) {
  if (is.factor(x)) x <- as.character(x)
  if (is.character(x)) sum(x == "Prefer not to answer", na.rm = TRUE) else 0
}))
```

#### Convert all chara vars to numeric

```{r, char-num}

#change pna to na in continuous variables
char_cols <- names(ukb_v2)[sapply(ukb_v2, is.character)]

#change character to numeric
ukb_v2[char_cols] <- lapply(ukb_v2[char_cols], function(x) {
  x[x == "Prefer not to answer"] <- NA
  suppressWarnings(as.numeric(x))
})

#final check
types_df <- data.frame(
  variable = names(ukb_v2),
  class = sapply(ukb_v2, function(x) paste(class(x), collapse = "/")),
  stringsAsFactors = FALSE
)

```

## 4) Reclassify ethnicity

```{r}
eth <- as.character(ukb_v2$ethnicity.0.0)
eth <- trimws(eth)


white_set <- c(
  "White",
  "British", "Irish", "Any other white background"
)

asian_set <- c(
  "Asian or Asian British",
  "Indian", "Pakistani", "Bangladeshi",
  "Any other Asian background",
  "Chinese"
)

mixed_set <- c(
  "Mixed",
  "White and Black Caribbean", "White and Black African", "White and Asian",
  "Any other mixed background"
)

black_set <- c("Black or Black British", "Caribbean", "African", "Any other Black background")

other_set <- c('Other ethnic group')


ukb_v2$ethnicity.0.0 <- dplyr::case_when(
  is.na(eth) ~ NA_character_,
  eth %in% white_set ~ "White",
  eth %in% asian_set ~ "Asian",
  eth %in% mixed_set ~ "Mixed",
  eth %in% black_set ~ "Black",
  eth %in% other_set ~ "Other")

ukb_v2$ethnicity.0.0  <- as.factor(ukb_v2$ethnicity.0.0)
```

## 5) Outlier removal


```{r, outliers}

ukb_v2 <- ukb_v2 %>%
  mutate(across(
    where(is.numeric),
    ~ ifelse(. < -1e5, NA, .)
  ))

```

## 6) Instance selection

Select the latest instance for all vars

```{r, instance}
# only have one date of death col (both cols are the exact same)

ukb_v2$date_death.0.0.1 <- NULL 

# keep latest ses_current_employment_status

ukb_v2 <- ukb_v2[, !grepl("ses_current_employment_status\\.0\\.[0-5]$", names(ukb_v2))]

# keep the latest ses_commute_transport_type

ukb_v2 <- ukb_v2[, !grepl("ses_commute_transport_type\\.0\\.[0-2]$", names(ukb_v2))]

# take the latest qualification col

ukb_v2 <- ukb_v2[, !grepl("qualification\\.0\\.[0-4]$", names(ukb_v2))]

```

## 7) Merging cols

```{r, merge}

# for reproducibility take automated sbp and dbp cols

# remove manual bp cols

ukb_v2 <- ukb_v2 %>% 
  dplyr::select(-sbp_manual.0.0, -sbp_manual.0.1, -dbp_manual.0.0, -dbp_manual.0.1)

# average sbp cols
ukb_v2 <- ukb_v2 %>%
  dplyr::mutate(
    dplyr::across(c(sbp_automated.0.0, sbp_automated.0.1), ~ {
      x <- as.numeric(as.character(.))
      x[x == 8242] <- NA
      x
    }),
    sbp_avg = rowMeans(dplyr::across(c(sbp_automated.0.0, sbp_automated.0.1)), na.rm = TRUE)
  ) %>%
  dplyr::select(-sbp_automated.0.0, -sbp_automated.0.1)

# average dbp cols
ukb_v2 <- ukb_v2 %>% 
  dplyr::mutate(
    dbp_avg = rowMeans(dplyr::across(c(dbp_automated.0.0, dbp_automated.0.1)), na.rm = TRUE)
  ) %>%
  dplyr::select(-dbp_automated.0.0, -dbp_automated.0.1)

# update domestic_garden_pct_buffer_300m.0.0 to start with env

ukb_v2$env_domestic_garden_pct_buffer_300m <- ukb_v2$domestic_garden_pct_buffer_300m.0.0

```


## 8) IMD tertiles

```{r, imd}

process_imd <- function(df) {
  
  # Map assessment centre to country
  
  england_centres <- c(
    "Stockport (pilot)", "Manchester", "Oxford", "Stoke", "Reading",
    "Bury", "Newcastle", "Leeds", "Bristol", "Barts", "Nottingham",
    "Sheffield", "Liverpool", "Middlesborough", "Hounslow",
    "Croydon", "Birmingham", "Cheadle (revisit)", "Cheadle (imaging)",
    "Reading (imaging)", "Newcastle (imaging)", "Bristol (imaging)"
  )
  
  scotland_centres <- c("Glasgow", "Edinburgh")
  
  wales_centres <- c("Cardiff", "Swansea", "Wrexham")
  
  df <- df %>%
    mutate(
      uk_country = case_when(
        assessment_centre_location.0.0 %in% england_centres  ~ "England",
        assessment_centre_location.0.0 %in% scotland_centres ~ "Scotland",
        assessment_centre_location.0.0 %in% wales_centres    ~ "Wales",
        TRUE ~ NA_character_
      )
    )
  
  # Country specific IMD vars
  
  df <- df %>%
    mutate(
      imd_raw = case_when(
        uk_country == "England"  ~ ses_imd_england.0.0,
        uk_country == "Scotland" ~ ses_imd_scotland.0.0,
        uk_country == "Wales"    ~ ses_imd_wales.0.0,
        TRUE ~ NA_real_
      )
    )
  
  # Country specific IMD tertiles
  
  df <- df %>%
    group_by(uk_country) %>%
    mutate(
      imd_tertile = ntile(imd_raw, 3)
    ) %>%
    ungroup()
  
  # Merge IMD var
  
  df <- df %>%
    mutate(
      ses_imd = imd_tertile
    )
  
  df <- df%>%
    dplyr::select(-uk_country, -imd_raw, -imd_tertile, -ses_imd_england.0.0, -ses_imd_scotland.0.0, -ses_imd_wales.0.0, -assessment_centre_location.0.0)
  
  return(df)
}

# remove assessment centre location (no longer needed)

ukb_v2 <- process_imd(ukb_v2)

```

## 9) Alcohol cats

```{r, cats}
process_alcohol <- function(df) {
  
  df %>%
    dplyr::mutate(
      
      #aclohol categories
      alcohol_category = case_when(
        
        # Never drinkers
        alc_freq.0.0 == "Never" ~ "Non-drinker",
        alc_status.0.0 == "Never" ~ "Non-drinker",
        
        # Former drinkers
        alc_status.0.0 == "Previous" ~ "Former",
        
        # Current drinkers by frequency
        alc_status.0.0 == "Current" &
          alc_freq.0.0 %in% c("Special occasions only",
                              "One to three times a month") ~ "Social drinker",
        
        alc_status.0.0 == "Current" &
          alc_freq.0.0 == "Once or twice a week" ~ "Moderate drinker",
        
        alc_status.0.0 == "Current" &
          alc_freq.0.0 %in% c("Three or four times a week",
                              "Daily or almost daily") ~ "Daily drinker",
        
        TRUE ~ NA_character_
      )
    )
}  


ukb_v2 <- process_alcohol(ukb_v2)
```

## 10) Missingness checks

```{r, missingness}

# column-wise missingness

column_missing_rate <- sapply(ukb_v2, function(x) mean(is.na(x)))

column_missing_summary <- data.frame(
  var = names(column_missing_rate),
  column_missing_rate = as.numeric(column_missing_rate)
)

# variables with >50% missingness
high_missing <- column_missing_summary %>%
  filter(column_missing_rate > 0.5) %>%
  mutate(
    percent_missing = round(column_missing_rate * 100, 1)
  ) %>%
  arrange(desc(percent_missing))

# view in R
high_missing

# save as csv
write.csv(
  high_missing,
  file = file.path(output_dir, "vars_missing_gt_50_percent.csv"),
  row.names = FALSE
)

# save just variable names as txt
writeLines(
  high_missing$var,
  con = file.path(output_dir, "vars_missing_gt_50_percent.txt")
)


# row-wise missingness

row_missing_rate <- apply(ukb_v2, 1, function(x) mean(is.na(x)))
row_missing_n    <- apply(ukb_v2, 1, function(x) sum(is.na(x)))

row_missing_summary <- data.frame(
  missing_n = row_missing_n,
  missing_rate = row_missing_rate
)


# summaries

summary(column_missing_rate)
summary(row_missing_rate)


# save histograms

png(file.path(output_dir, "column_missingness_hist.png"), width = 1800, height = 1200, res = 200)
hist(
  column_missing_rate,
  breaks = 50,
  main = "Column-wise missingness",
  xlab = "Missing rate"
)
dev.off()

png(file.path(output_dir, "row_missingness_hist.png"), width = 1800, height = 1200, res = 200)
hist(
  row_missing_rate,
  breaks = 50,
  main = "Row-wise missingness",
  xlab = "Missing rate"
)
dev.off()

# hist
hist(column_missing_rate, breaks = 50, main = "Column-wise missingness", xlab = "Missing rate")
hist(row_missing_rate, breaks = 50, main = "Row-wise missingness", xlab = "Missing rate")
```

#### Domain level missingness


```{r}

# 1. Get variable names
var_names <- names(ukb_v2)

# 2. Assign each variable to a domain based on its prefix/name

# Notes:
# - startsWith() is used for prefixes
# - exact variable names are matched with %in%
# - variables that do not match any rule are left as NA

domain_map <- tibble(variable = var_names) %>%
  mutate(
    domain = case_when(
      # Biological
      startsWith(variable, "biochem_") | startsWith(variable, "blood_") ~ "Biological",
      
      # Lifestyle
      startsWith(variable, "alc") |
        startsWith(variable, "diet_") |
        startsWith(variable, "phys_") |
        startsWith(variable, "sleep_") |
        startsWith(variable, "smoke_") |
        startsWith(variable, "sed_") ~ "Lifestyle",

      # Environmental
      startsWith(variable, "env_") ~ "Environmental",
      
      # Socioeconomic
      startsWith(variable, "ses_") |
        startsWith(variable, "qualification") ~ "Socioeconomic",
      
      # Psychosocial
      startsWith(variable, "mh_") ~ "Psychosocial",
      
      # Baseline
      variable %in% c("ethnicity.0.0", "age_recr.0.0", "sex.0.0") |
        startsWith(variable, "sbp_") |
        startsWith(variable, "dbp_") |
        startsWith(variable, "bmi") ~ "Baseline",
      
      TRUE ~ NA_character_
    )
  )

# Optional: inspect variables not assigned to any domain
unassigned_vars <- domain_map %>%
  filter(is.na(domain))

print(unassigned_vars)

# 3. Calculate % missing for each variable
var_missing <- tibble(
  variable = names(ukb_v2),
  pct_missing = sapply(ukb_v2, function(x) mean(is.na(x)) * 100)
)

# 4. Merge domain assignment with missingness values
domain_missing <- domain_map %>%
  left_join(var_missing, by = "variable") %>%
  filter(!is.na(domain))

# 5. Calculate mean % missing within each domain
mean_domain_missing <- domain_missing %>%
  group_by(domain) %>%
  summarise(
    mean_pct_missing = mean(pct_missing, na.rm = TRUE),
    n_variables = n(),
    .groups = "drop"
  )

print(mean_domain_missing)

# 6. Order domains from highest to lowest missingness
mean_domain_missing <- mean_domain_missing %>%
  arrange(desc(mean_pct_missing)) %>%
  mutate(domain = factor(domain, levels = domain))

# 7. Plot

p <- ggplot(mean_domain_missing, aes(x = mean_pct_missing, y = domain)) +
  geom_col(fill = "#1f3b73", colour = "grey30") +
  labs(
    title = "Mean Missingness by Domain",
    x = "Mean % Missing",
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

p

ggsave(
  "../pipeline_outputs/pre_output/mean_missingness_by_domain.png",
  plot = p,
  width = 8,
  height = 5,
  dpi = 300
)

```


## 11) Excluding vars

Update this based on which vars have a missingness > 50% in ukbiobank

Vars have been excluded based on their level of missingness in the UK Biobank. Vars with missingness >50% are excluded.

```{r, exc}

ukb_v2 <- ukb_v2%>%
  dplyr::select( -starts_with("smoke_age"), 
         -starts_with("smoke_tobacco_type_"), -diet_carb.0.0,
         -sleep_wake_early.0.0, -sleep_too_much.0.0, 
         -sleep_falling_asleep_trouble.0.0, -sleep_change.0.0, -sleep_daytime_nap.0.0, 
         -smoke_people_in_household.0.0, -smoke_tobacco_exposure_home.0.0,  
         -smoke_tobacco_exposure_outside.0.0)

```


## 12) Table 1
```{r}
# ---- helper functions ----
fmt_mean_sd <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  paste0(round(mean(x), digits), " (", round(sd(x), digits), ")")
}

fmt_n <- function(x) {
  sum(!is.na(x))
}

cat_table <- function(x, varname, digits = 1, include_na = FALSE) {
  if (!include_na) {
    tab <- table(x, useNA = "no")
    denom <- sum(tab)
  } else {
    tab <- table(x, useNA = "ifany")
    denom <- sum(tab)
  }
  prop <- (tab / denom) * 100
  
  data.frame(
    Variable = varname,
    Level    = names(tab),
    Summary  = paste0(as.integer(tab), " (", round(as.numeric(prop), digits), "%)"),
    stringsAsFactors = FALSE
  )
}

# ---- specify variables ----
cont_vars <- c("age_recr.0.0", "sbp_avg", "dbp_avg", "mh_neuroticism_score.0.0", "bmi.0.0")

cat_vars <- c("sex.0.0",
              "ethnicity.0.0",
              "cvd_event",
              "smoke_status.0.0",
              "alc_status.0.0",
              "alc_category",
              "ses_current_employment_status.0.6",
              "ses_imd",
              "qualification.0.5",
              "phys_ipaq.0.0")

# ---- build table 1 ----
table1_list <- list()

# add overall N
table1_list[["N_total"]] <- data.frame(
  Variable = "Participants",
  Level = "",
  Summary = as.character(nrow(ukb_v2)),
  stringsAsFactors = FALSE
)

# continuous vars: Mean (SD)
for (v in cont_vars) {
  x <- ukb_v2[[v]]
  table1_list[[v]] <- data.frame(
    Variable = v,
    Level = "",
    Summary = paste0(fmt_mean_sd(x), "; n=", fmt_n(x)),
    stringsAsFactors = FALSE
  )
}

# categorical vars: n (%)
for (v in cat_vars) {
  x <- ukb_v2[[v]]
  table1_list[[v]] <- cat_table(x, varname = v, digits = 1, include_na = FALSE)
}

table1_df <- do.call(rbind, table1_list)

# view in console
print(table1_df, row.names = FALSE)

# export
write.csv(table1_df, file.path(output_dir, "Table1_baseR.csv"), row.names = FALSE)
```
#### Table 1 version 2 (using tableone function)

```{r, table1}

# relable IMD tertiles to improve interpretability

ukb_v2$ses_imd <- factor(
  ukb_v2$ses_imd,
  levels = c(1, 2, 3),
  labels = c("Least deprived", "Middle", "Most deprived"),
  ordered = TRUE
)

# update var names to be more interpretable

ukb_table1 <- ukb_v2 %>%
  rename(
    `Age at recruitment, years` = age_recr.0.0,
    `Systolic blood pressure, mmHg` = sbp_avg,
    `Diastolic blood pressure, mmHg` = dbp_avg,
    `Neuroticism score` = mh_neuroticism_score.0.0,
    `Sex` = sex.0.0,
    `Ethnicity` = ethnicity.0.0,
    `BMI` = bmi.0.0,
    `Smoking status` = smoke_status.0.0,
    `Alcohol drinking status` = alc_status.0.0,
    `Alcohol category` = alc_category,
    `Employment status` = ses_current_employment_status.0.6,
    `Index of multiple deprivation tertile` = ses_imd,
    `Educational qualification` = qualification.0.5,
    `Physical activity status (IPAQ)` = phys_ipaq.0.0
  )

# make stratifying variable readable 
ukb_table1$cvd_event <- factor(
  ukb_table1$cvd_event,
  levels = c(0, 1),
  labels = c("No incident CVD", "Incident CVD")
)

# variables to include 
cont_vars <- c(
  "Age at recruitment, years",
  "Systolic blood pressure, mmHg",
  "Diastolic blood pressure, mmHg",
  "Neuroticism score",
  "BMI"
)

cat_vars <- c(
  "Sex",
  "Ethnicity",
  "Smoking status",
  "Alcohol drinking status",
  "Alcohol category",
  "Employment status",
  "Index of multiple deprivation tertile",
  "Educational qualification",
  "Physical activity status (IPAQ)"
)

table1_vars <- c(cont_vars, cat_vars)

# keep only variables that exist
table1_vars <- table1_vars[table1_vars %in% names(ukb_table1)]
cat_vars <- cat_vars[cat_vars %in% names(ukb_table1)]

# create table 1 with tableone 
table1_obj <- CreateTableOne(
  vars = table1_vars,
  strata = "cvd_event",
  data = ukb_table1,
  factorVars = cat_vars,
  includeNA = FALSE
)

# print in console
print(table1_obj, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE)

# export as csv
table1_df <- print(
  table1_obj,
  showAllLevels = TRUE,
  quote = FALSE,
  noSpaces = TRUE,
  printToggle = FALSE
)

table1_df <- data.frame(
  Variable = rownames(table1_df),
  table1_df,
  row.names = NULL,
  check.names = FALSE
)

write.csv(
  table1_df,
  file.path(output_dir, "Table1_tableone_cvd_event.csv"),
  row.names = FALSE
)

# optional txt export
capture.output(
  print(table1_obj, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE),
  file = file.path(output_dir, "Table1_tableone_cvd_event.txt")
)


```



## 13) Imputation

For imputation, run the corresponding imputation script in `pipeline_scripts/`.

This step splits the dataset into training and test sets, fits the imputation model on the training data, and applies the trained model to the test data.

Imputed datasets are saved to the designated output directory.

##### Save

```{r, save}
saveRDS(ukb_v2, file = file.path(output_dir, "ukb_raw.rds"))

```
