---
title: "Preprocessing"
author: "Group_8"
date: "2026-03-01"
output: html_document
---

# Workflow overview:

0)  Setup
1)  Distribution checks
2)  Exclusion of individuals with CVD before baseline
3)  Conversion of "Do not know" (IDK) and "Prefer not to answer" (PNA) to NA
4)  Reclassify ethnicity
5)  Removal of outliers
6)  Instance selection
7)  Merging/dropping of variables
8)  Skip logic
9)  IMD tertiles
10) Missingness checks
11) Vars exclusion
12) Imputation

## 0) Setup

Load all necessary libraries

```{r, libraries}
library(dplyr)
library(lubridate)
library(stringr)
```

Load data

```{r, data}
ukb <- readRDS("../extraction_and_recoding/outputs/ukb_recoded.rds")
cvd_event <- readRDS("cvd_events.rds")
```

Addition of CVD_event, CVD_date and EID cols to data

```{r, prelim}

# add eid column
ukb$eid <- rownames(ukb)
rownames(ukb) <- NULL
ukb$eid <- as.character(ukb$eid)

cvd_event$eid <- as.character(cvd_event$eid)

# cvd event flag
cvd_ids <- unique(cvd_event$eid)
ukb$cvd_event <- as.integer(ukb$eid %in% cvd_ids)

# first cvd date 
first_cvd <- aggregate(date ~ eid, cvd_event, min)

ukb$cvd_date <- first_cvd$date[match(ukb$eid, first_cvd$eid)]

# reorder columns
first_cols <- c("eid", "cvd_event", "cvd_date")
other_cols <- setdiff(names(ukb), first_cols)
other_cols_sorted <- sort(other_cols)
new_order <- c(first_cols, other_cols_sorted)
ukb <- ukb[, new_order]

# check binary
table(ukb$cvd_event, useNA = "ifany")
```

## 1) Distribution checks

Distribution checks are ran on the dataset before any changes. /rds/general/project/hda_25-26/live/TDS/TDS_Group8/preprocessing/distribution_check/dis_full.html - contains all the distribution checks of this dataset at baseline.

## 2) Exclusion

Exclude individuals with CVD_event before recruitment

*Included a 1 year wash out period*

```{r, exclusion}

ukb_v2 <- ukb %>%
  mutate(
    date_recr.0.0 = as.Date(date_recr.0.0),
    cvd_date      = as.Date(cvd_date),
    
    cvd_case = case_when(
      !is.na(cvd_date) &
        cvd_date <= date_recr.0.0 + years(1) ~ "exclude_washout",
      
      !is.na(cvd_date) &
        cvd_date > date_recr.0.0 + years(1) ~ "incident",
      
      TRUE ~ "no_event"
    )
  )

table(ukb_v2$cvd_case)

# Exclude washout cases
ukb_v2 <- ukb_v2 %>%
  filter(cvd_case != "exclude_washout") %>%
  dplyr::select(-cvd_case)

table(ukb_v2$cvd_event)
```

## 3) PNA & IDK to NA conversion

Convert all O First check the number of PNA and IDKs throughout the entire dataset.

```{r, check}
N <- nrow(ukb_v2)

# Helper function to compute rates
compute_rate <- function(df, target_string) {
  # Rate among non-NA
  rate_nonNA <- sapply(df, function(x) {
    if (is.character(x) || is.factor(x)) {
      mean(as.character(x) == target_string, na.rm = TRUE)
    } else NA_real_
  })
  
  df_nonNA <- data.frame(
    column = names(rate_nonNA),
    rate_nonNA = as.numeric(rate_nonNA),
    row.names = NULL
  )
  
  # Rate over total rows
  rate_total <- sapply(df, function(x) {
    if (is.character(x) || is.factor(x)) {
      sum(as.character(x) == target_string, na.rm = TRUE) / N
    } else NA_real_
  })
  
  df_total <- data.frame(
    column = names(rate_total),
    rate_total = as.numeric(rate_total),
    row.names = NULL
  )
  
  merged <- merge(
    df_total,
    df_nonNA,
    by = "column",
    all = TRUE,
    suffixes = c("_overall", "_nonNA")
  )
  
  names(merged)[names(merged) == "rate_total"] <- paste0(target_string, "_rate_overallN")
  names(merged)[names(merged) == "rate_nonNA"] <- paste0(target_string, "_rate_among_nonmissing")
  
  merged
}

# Compute PNA rates
pna_table <- compute_rate(ukb_v2, "Prefer not to answer")

# Compute IDK rates
idk_table <- compute_rate(ukb_v2, "I don't know")

# Merge both PNA and IDK tables
merged_table <- merge(pna_table, idk_table, by = "column", all = TRUE)

# Order by highest overall PNA rate (optional)
merged_table <- merged_table[order(-merged_table$`Prefer not to answer_rate_overallN`), ]

# View result
head(merged_table)

```

#Convert IDK and PNA to NAs

```{r, NA}
#change all IDK to NAs
ukb_v2[] <- lapply(ukb_v2, function(x) {
  if (is.character(x) || is.factor(x)) {
    x[x == "Do not know"] <- NA
  }
  x
})

#check
sum(sapply(ukb_v2, function(x) {
  if (is.factor(x)) x <- as.character(x)
  if (is.character(x)) sum(x == "Do not know", na.rm = TRUE) else 0
}))


# Convert all "Prefer not to answer" to NA
ukb_v2[] <- lapply(ukb_v2, function(x) {
  if (is.character(x) || is.factor(x)) {
    x[x == "Prefer not to answer"] <- NA
  }
  x
})

# Quick check: should be 0 now
sum(sapply(ukb_v2, function(x) {
  if (is.factor(x)) x <- as.character(x)
  if (is.character(x)) sum(x == "Prefer not to answer", na.rm = TRUE) else 0
}))
```

#Convert all chara vars to numeric

```{r, char-num}

#change pna to na in continuous variables
char_cols <- names(ukb_v2)[sapply(ukb_v2, is.character)]

#change character to numeric
ukb_v2[char_cols] <- lapply(ukb_v2[char_cols], function(x) {
  x[x == "Prefer not to answer"] <- NA
  suppressWarnings(as.numeric(x))
})

#final check
types_df <- data.frame(
  variable = names(ukb_v2),
  class = sapply(ukb_v2, function(x) paste(class(x), collapse = "/")),
  stringsAsFactors = FALSE
)

```

## 4) reclassify ethnicity

```{r}
eth <- as.character(ukb_v2$ethnicity.0.0)
eth <- trimws(eth)


white_set <- c(
  "White",
  "British", "Irish", "Any other white background"
)

asian_set <- c(
  "Asian or Asian British",
  "Indian", "Pakistani", "Bangladeshi",
  "Any other Asian background",
  "Chinese"
)

mixed_set <- c(
  "Mixed",
  "White and Black Caribbean", "White and Black African", "White and Asian",
  "Any other mixed background"
)

black_set <- c("Black or Black British", "Caribbean", "African", "Any other Black background")

other_set <- c('Other ethnic group')


ukb_v2$ethnicity.0.0 <- dplyr::case_when(
  is.na(eth) ~ NA_character_,
  eth %in% white_set ~ "White",
  eth %in% asian_set ~ "Asian",
  eth %in% mixed_set ~ "Mixed",
  eth %in% black_set ~ "Black",
  eth %in% other_set ~ "Other")

ukb_v2$ethnicity.0.0  <- as.factor(ukb_v2$ethnicity.0.0)
```

## 5) Outlier removal

*For ukbiobank edit this so all values "\< 0" becomes NA*

```{r, outliers}

ukb_v2 <- ukb_v2 %>%
  mutate(across(
    where(is.numeric),
    ~ ifelse(. < -1e5, NA, .)
  ))

```

## 6) Instance selection

Select the latest instance for all vars

*For ukbiobank need to check how instances are ordered compared with synthetic, i.e., is each col one category?*

```{r, instance}
# only have one date of death col (both cols are the exact same)

ukb_v2$date_death.0.0.1 <- NULL 

# keep latest ses_current_employment_status

ukb_v2 <- ukb_v2[, !grepl("ses_current_employment_status\\.0\\.[0-5]$", names(ukb_v2))]

# keep the latest ses_commute_transport_type

ukb_v2 <- ukb_v2[, !grepl("ses_commute_transport_type\\.0\\.[0-2]$", names(ukb_v2))]

# take the latest qualification col

ukb_v2 <- ukb_v2[, !grepl("qualification\\.0\\.[0-4]$", names(ukb_v2))]

```

## 7) Merging cols

```{r, merge}

# for reproducibility take automated sbp and dbp cols

# remove manual bp cols

ukb_v2 <- ukb_v2 %>% 
  dplyr::select(-sbp_manual.0.0, -sbp_manual.0.1, -dbp_manual.0.0, -dbp_manual.0.1)

# average sbp cols
ukb_v2 <- ukb_v2 %>%
  dplyr::mutate(
    dplyr::across(c(sbp_automated.0.0, sbp_automated.0.1), ~ {
      x <- as.numeric(as.character(.))
      x[x == 8242] <- NA
      x
    }),
    sbp_avg = rowMeans(dplyr::across(c(sbp_automated.0.0, sbp_automated.0.1)), na.rm = TRUE)
  ) %>%
  dplyr::select(-sbp_automated.0.0, -sbp_automated.0.1)

# average dbp cols
ukb_v2 <- ukb_v2 %>% 
  dplyr::mutate(
    dbp_avg = rowMeans(dplyr::across(c(dbp_automated.0.0, dbp_automated.0.1)), na.rm = TRUE)
  ) %>%
  dplyr::select(-dbp_automated.0.0, -dbp_automated.0.1)

# average fev1 and fvc cols
ukb_v2 <- ukb_v2 %>%
  dplyr::mutate(
    fev_avg = rowMeans(dplyr::across(c(fev1.0.0, fev1.0.1, fev1.0.2)), na.rm = TRUE),
    dplyr::across(c(fvc.0.0, fvc.0.1, fvc.0.2), ~ as.numeric(as.character(.))),
    fvc_avg = rowMeans(dplyr::across(c(fvc.0.0, fvc.0.1, fvc.0.2)), na.rm = TRUE)
  ) %>%
  dplyr::select(-fvc.0.0, -fvc.0.1, -fvc.0.2, -fev1.0.0, -fev1.0.1, -fev1.0.2)

# drop data
ukb_v2 <- ukb_v2 %>%
  dplyr::select(
    -sleep_snore.1.0, -sleep_snore.2.0, -sleep_snore.3.0,
    -sleep_chronotype.1.0, -sleep_chronotype.2.0, -sleep_chronotype.3.0
  )

# remove synthetic mob and yob <- DO NOT DO THIS IN BIOBANK

ukb_v2 <- ukb_v2%>% 
  dplyr::select(-yob.0.0, -mob.0.0)

# Extra date of death col needs to be removed

# ukb_v2 <- ukb_v2%>% 
#  select(-cvd_date.1)

```

## 8) Skip logic

Need to ask whether we should be doing this before/after imputation. For now I made two datasets. One that deals with skip logic beforehand, and one that deals with it after imputation.

#### Smoke skip logic

```{r, smoke skip logic}

ukb_v3 <- ukb_v2

current_smoker <- ukb_v3$smoke_tobacco_current.0.0 =="Yes, on most or all days"
not_current_smoker <- !current_smoker
ukb_v3$smoke_current_indicator <- ifelse(current_smoker, 1, 0)

ukb_v3$smoke_tobacco_type_current.0.0 <- as.character(ukb_v3$smoke_tobacco_type_current.0.0)
ukb_v3$smoke_tobacco_type_current.0.0[not_current_smoker] <- "None"
ukb_v3$smoke_number_daily_current_smoker.0.0[not_current_smoker] <- 0


# field 1249
past_smoker <- ukb_v3$smoke_tobacco_past.0.0 == 'Smoked on most or all days'
not_past_smoker <- !past_smoker
ukb_v3$smoke_past_indicator <- ifelse(past_smoker, 1, 0)

ukb_v3$smoke_tobacco_type_previous.0.0 <- as.character(ukb_v3$smoke_tobacco_type_previous.0.0)
ukb_v3$smoke_tobacco_type_previous.0.0[not_past_smoker] <- 'None'
ukb_v3$smoke_cig_daily_previous.0.0[not_past_smoker] <- 0


# field 3446
cigar_pipe <- ukb_v3$smoke_tobacco_type_current.0.0 == "Cigars or pipes"
not_cigar_pipe <- !cigar_pipe 
ukb_v3$cigar_pipe_indicator <- ifelse(cigar_pipe, 1, 0)
ukb_v3$smoke_number_cig_previous_current_cigar_pipe.0.0[not_cigar_pipe] <- 0
```

#### Diet skip logic

*Removed coffee intake due to it not being available on UKBiobank*

```{r}
##coffee-------------------
#add "None" as answer to "diet_coffee_type.0.0" if "diet_coffee_intake" = 0
ukb_v3$diet_coffee_type.0.0 <- as.character(ukb_v3$diet_coffee_type.0.0)
# ukb_v3$diet_coffee_type.0.0[ukb_v3$diet_coffee_intake.0.0 == 0] <- "None"

##bread----------------------------
#add "None" as answer to "diet_bread_type.0.0" if "diet_bread_intake" = 0
ukb_v3$diet_bread_type.0.0 <- as.character(ukb_v3$diet_bread_type.0.0)
ukb_v3$diet_bread_type.0.0[ukb_v3$diet_bread_intake.0.0 == 0] <- "None"

#cereal----------------------------
#add "None" as answer to "diet_cereal_type.0.0" if "diet_cereal_intake" = 0
ukb_v3$diet_cereal_type.0.0 <- as.character(ukb_v3$diet_cereal_type.0.0)
ukb_v3$diet_cereal_type.0.0[ukb_v3$diet_cereal_intake.0.0 == 0] <- "None"

#alc_wih_meals---------------------
#add "None" as answer to "alc_with_meals.0.0" if "alc_freq" = "Never"
ukb_v3$alc_freq.0.0 <- as.character(ukb_v3$alc_freq.0.0)
ukb_v3$alc_with_meals.0.0 <- as.character(ukb_v3$alc_with_meals.0.0)
ukb_v3$alc_with_meals.0.0[ukb_v3$alc_freq.0.0 == "Never"] <- "None"
```

#### Medication skip logic

```{r}
# med_chol_bp_diabetes.0\.\0-2 <- male
# med_chol_bp_diabetes_hormones.0\.\1-3 <- female

# base off sex.0.0

med_male <- grep("^med_chol_bp_diabetes\\.0\\.[0-2]+$",
                             names(ukb_v3),
                             value = TRUE)

med_female <- grep("^med_chol_bp_diabetes_hormones\\.0\\.[0-3]+$",
                               names(ukb_v3),
                               value = TRUE)

# Male meds: convert to character first, then apply NA for females
ukb_v3 <- ukb_v3 %>%
  mutate(across(all_of(med_male),
                ~ ifelse(sex.0.0 == "Female", NA, as.character(.)))) %>%
  
  # Female meds: convert to character first, then apply NA for males
  mutate(across(all_of(med_female),
                ~ ifelse(sex.0.0 == "Male", NA, as.character(.))))
```

#### SES skip logic

```{r}
# create employed indicator 

ukb_v3 <- ukb_v3 %>%
  mutate(
    employed = ifelse(
      ses_current_employment_status.0.6 %in%
        c("In paid employment or self-employed"),
      1, 0
    )
  )

#creating chars
ukb_v3$ses_job_shift_work.0.0 <- as.character(ukb_v3$ses_job_shift_work.0.0)
ukb_v3$ses_job_mainly_walking_or_standing.0.0 <- as.character(ukb_v3$ses_job_mainly_walking_or_standing.0.0)
ukb_v3$ses_job_heavy_manual_or_physical_work.0.0 <- as.character(ukb_v3$ses_job_heavy_manual_or_physical_work.0.0)
ukb_v3$ses_commute_transport_type.0.3 <- as.character(ukb_v3$ses_commute_transport_type.0.3)

# ses_length_working_week.0.0
ukb_v3$ses_length_working_week.0.0[ukb_v3$employed %in% c(0)] <- 0

# ses_job_shift_work.0.0
ukb_v3$ses_job_shift_work.0.0[ukb_v3$employed %in% c(0)] <- NA

# ses_job_mainly_walking_or_standing.0.0
ukb_v3$ses_job_mainly_walking_or_standing.0.0[ukb_v3$employed %in% c(0)] <- NA

# ses_job_heavy_manual_or_physical_work
ukb_v3$ses_job_heavy_manual_or_physical_work.0.0[ukb_v3$employed %in% c(0)] <- NA

# ses_commute_transport_type
ukb_v3$ses_commute_transport_type.0.3[ukb_v3$employed %in% c(0)] <- NA

# ses_commute_frequency
ukb_v3$ses_commute_frequency.0.0[ukb_v3$employed %in% c(0)] <- 0

# ses_commute_distance_home_to_work
ukb_v3$ses_commute_distance_home_to_work.0.0[ukb_v3$employed %in% c(0)] <- 0

# ses_age_completed_full_time_education
ukb_v3$ses_age_completed_full_time_education.0.0[ukb_v3$qualification.0.5 %in% c("College or University degree")] <- NA

# ses_household_income_before_tax_avg
ukb_v3$ses_household_income_before_tax_avg.0.0[ukb_v3$ses_accom_type.0.0 %in% c("Sheltered accommodation", "Care home")] <- NA

```

## 9) IMD tertiles

```{r, imd}

process_imd <- function(df) {
  
  # Map assessment centre to country
  
  england_centres <- c(
    "Stockport (pilot)", "Manchester", "Oxford", "Stoke", "Reading",
    "Bury", "Newcastle", "Leeds", "Bristol", "Barts", "Nottingham",
    "Sheffield", "Liverpool", "Middlesborough", "Hounslow",
    "Croydon", "Birmingham", "Cheadle (revisit)", "Cheadle (imaging)",
    "Reading (imaging)", "Newcastle (imaging)", "Bristol (imaging)"
  )
  
  scotland_centres <- c("Glasgow", "Edinburgh")
  
  wales_centres <- c("Cardiff", "Swansea", "Wrexham")
  
  df <- df %>%
    mutate(
      uk_country = case_when(
        assessment_centre_location.0.0 %in% england_centres  ~ "England",
        assessment_centre_location.0.0 %in% scotland_centres ~ "Scotland",
        assessment_centre_location.0.0 %in% wales_centres    ~ "Wales",
        TRUE ~ NA_character_
      )
    )
  
  # Country specific IMD vars
  
  df <- df %>%
    mutate(
      imd_raw = case_when(
        uk_country == "England"  ~ ses_imd_england.0.0,
        uk_country == "Scotland" ~ ses_imd_scotland.0.0,
        uk_country == "Wales"    ~ ses_imd_wales.0.0,
        TRUE ~ NA_real_
      )
    )
  
  # Country specific IMD tertiles
  
  df <- df %>%
    group_by(uk_country) %>%
    mutate(
      imd_tertile = ntile(imd_raw, 3)
    ) %>%
    ungroup()
  
  # Merge IMD var
  
  df <- df %>%
    mutate(
      imd = imd_tertile
    )
  
  df <- df%>%
    dplyr::select(-uk_country, -imd_raw, -imd_tertile, -ses_imd_england.0.0, -ses_imd_scotland.0.0, -ses_imd_wales.0.0)
  
  return(df)
}

ukb_v2 <- process_imd(ukb_v2)
ukb_v3 <- process_imd(ukb_v3)

```

## 10) Missingness checks

```{r, missingness}

# Check column-wise missingness

column_missing_rate <- sapply(ukb_v3, function(x) mean(is.na(x)))

column_missing_summary <- data.frame(
  var = names(ukb_v3),
  column_missing_rate = column_missing_rate
)

# pull high missing vars 

high_missing <- column_missing_summary %>%
  filter(column_missing_rate > 0.5) %>%
  mutate(
    percent_missing = round(column_missing_rate * 100, 1)
  ) %>%
  arrange(desc(percent_missing))

high_missing
# Check row-wise missingness

exclude_vars <- c("cvd_date", 
                  "date_death.0.0", 
                  "date_lost_to_fup.0.0")

# Create subset excluding those columns
ukb_subset <- ukb_v3[, !(names(ukb_v3) %in% exclude_vars)]

# Row-wise missing rate
row_missing_rate <- apply(ukb_subset, 1, function(x) mean(is.na(x)))

# Optional: number of missing per row
row_missing_n <- apply(ukb_subset, 1, function(x) sum(is.na(x)))

# Combine into summary dataframe
row_missing_summary <- data.frame(
  missing_n = row_missing_n,
  missing_rate = row_missing_rate
)

# visualising missingness
# column summary
summary(column_missing_rate)
hist(column_missing_rate, breaks = 50)

# row summary
summary(row_missing_rate)
hist(row_missing_rate, breaks = 50)

```

## 11) Excluding vars

Vars have been excluded based on their level of missingness in the UK Biobank

*Summed MET minutes/week is unavailable on Biobank, other vars have \>50% NA*

```{r, exc}

ukb_v3 <- ukb_v3%>%
  dplyr::select(-phys_MET_week.0.0, -starts_with("smoke_age"), 
         -starts_with("smoke_tobacco_type_"), -diet_carb.0.0, -diet_energy.0.0, -diet_fat.0.0,
         -diet_food_weight.0.0, -diet_polyunsat_fat.0.0, -diet_protein.0.0, -diet_sat_fat.0.0,
         -diet_total_sugar.0.0, -sleep_wake_early.0.0, -sleep_too_much.0.0, 
         -sleep_falling_asleep_trouble.0.0, -sleep_change.0.0, -sleep_daytime_nap.0.0, 
         -smoke_people_in_household.0.0, -smoke_tobacco_exposure_home.0.0,  
         -smoke_tobacco_exposure_outside.0.0)

ukb_v2 <- ukb_v2%>%
  dplyr::select(-phys_MET_week.0.0, -starts_with("smoke_age"), 
         -starts_with("smoke_tobacco_type_"), -diet_carb.0.0, -diet_energy.0.0, -diet_fat.0.0,
         -diet_food_weight.0.0, -diet_polyunsat_fat.0.0, -diet_protein.0.0, -diet_sat_fat.0.0,
         -diet_total_sugar.0.0, -sleep_wake_early.0.0, -sleep_too_much.0.0, 
         -sleep_falling_asleep_trouble.0.0, -sleep_change.0.0, -sleep_daytime_nap.0.0, 
         -smoke_people_in_household.0.0, -smoke_tobacco_exposure_home.0.0,  
         -smoke_tobacco_exposure_outside.0.0)

```

## 12) Imputation

For imputation, run the corresponding imputation script in `pipeline_scripts/`.

This step splits the dataset into training and test sets, fits the imputation model on the training data, and applies the trained model to the test data.

Imputed datasets are saved to the designated output directory.

##### Save

```{r, save}
output_dir <- "../pipeline_outputs"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

saveRDS(ukb_v2, file = file.path(output_dir, "ukb_raw.rds"))
saveRDS(ukb_v3, file = file.path(output_dir, "ukb_skip.rds"))

```
