---
title: "Post Imputation Script"
author: "Group_8"
date: "2026-03-01"
output: html_document
---

------------------------------------------------------------------------

# Workflow overview:

0)  Setup

Part A) Exploratory analysis

1)  Distribution checks
2)  Dealing with skip logic
3)  Table 1
4)  Manhattan plot
5)  External exposome correlation analysis

Part B) Data finalisation

6)  One hot encoding of categorical vars
7)  Score creation
8)  Categories
9)  External exposome specific dataset creation

## 0) Setup

Load all necessary libraries

```{r setup, libraries}
library(dplyr)
library(stringr)
library(corrplot)
library(ggplot2)
library(GGally)
library(data.table)
library(broom)
```

Load datasets

```{r, data}
ukb_train <- readRDS("../pipeline_outputs/imputation_output/ukb_train.rds")
ukb_test  <- readRDS("../pipeline_outputs/imputation_output/ukb_test.rds")
ukb_raw   <- readRDS("../pipeline_outputs/pre_output/ukb_raw.rds")

output_dir <- "../pipeline_outputs/post_imputation_output"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1) Distribution checks

### Check negative values

Detect negative values

```{r, dis}

neg_vars <- function(df) {
  num_cols <- names(df)[sapply(df, is.numeric)]
  num_cols[sapply(num_cols, function(v) any(df[[v]] < 0, na.rm = TRUE))]
}

neg_before_vars <- neg_vars(ukb_raw)
neg_after_vars  <- neg_vars(ukb_train)

only_after  <- setdiff(neg_after_vars,  neg_before_vars) # negatives introduced by imputation
only_after
```

### Check impossible or extreme values

```{r}
# Rewrite impossible environmental percentage values

env_pct_vars <- c(
  "env_domestic_garden_pct_buffer_300m.0.0",
  "env_greenspace_pct_buffer_300m.0.0",
  "env_natural_env_perc_300.0.0",
  "env_water_perc_300.0.0"
)

for(v in env_pct_vars){
  ukb_train[[v]] <- pmax(ukb_train[[v]], 0)
  ukb_test[[v]]  <- pmax(ukb_test[[v]], 0)
}

```

## 2) Skip logic

Ensure skip logic and structural NAs are being respected after imputation - will also one hot encode all categorical skip vars

### Smoke skip logic

```{r smoke}
process_smoking_vars <- function(df) {
  
  # Helper: create smoking indicator and set amount column to 0
  # for people who are explicitly not in the target smoking group
  add_smoking_indicator <- function(data, status_col, target_value, indicator_col, amount_col) {
    if (all(c(status_col, amount_col) %in% names(data))) {
      
      status <- data[[status_col]]
      
      # indicator: 1 = target smoker, 0 = explicitly not, NA = missing
      data[[indicator_col]] <- ifelse(
        is.na(status),
        NA,
        ifelse(status == target_value, 1, 0)
      )
      
      # set amount = 0 only for people explicitly not in target group
      data[[amount_col]][!is.na(status) & status != target_value] <- 0
    }
    return(data)
  }
  
  # Current smoker
  df <- add_smoking_indicator(
    data         = df,
    status_col   = "smoke_tobacco_current.0.0",
    target_value = "Yes, on most or all days",
    indicator_col = "smoke_current_indicator",
    amount_col   = "smoke_number_daily_current_smoker.0.0"
  )
  
  # Past smoker
  df <- add_smoking_indicator(
    data         = df,
    status_col   = "smoke_tobacco_past.0.0",
    target_value = "Smoked on most or all days",
    indicator_col = "smoke_past_indicator",
    amount_col   = "smoke_cig_daily_previous.0.0"
  )
  
  return(df)
}

ukb_train <- process_smoking_vars(ukb_train)
ukb_test  <- process_smoking_vars(ukb_test)
```

### Diet skip logic

```{r, diet}

process_diet_none <- function(df) {
  
  # Helper to safely modify a pair of columns
  add_none_if_zero <- function(data, type_col, intake_col) {
    if (all(c(type_col, intake_col) %in% names(data))) {
      data[[type_col]] <- as.character(data[[type_col]])
      data[[type_col]][data[[intake_col]] == 0] <- "None"
    }
    return(data)
  }
  
  # Bread
  df <- add_none_if_zero(
    df,
    type_col   = "diet_bread_type.0.0",
    intake_col = "diet_bread_intake.0.0"
  )
  
  # Cereal
  df <- add_none_if_zero(
    df,
    type_col   = "diet_cereal_type.0.0",
    intake_col = "diet_cereal_intake.0.0"
  )
  
  # Coffee (just convert to character if present)
  if ("diet_coffee_type.0.0" %in% names(df)) {
    df$diet_coffee_type.0.0 <- as.character(df$diet_coffee_type.0.0)
  }
  
  # Alcohol with meals
  if (all(c("alc_with_meals.0.0", "alc_freq.0.0") %in% names(df))) {
    df$alc_freq.0.0       <- as.character(df$alc_freq.0.0)
    df$alc_with_meals.0.0 <- as.character(df$alc_with_meals.0.0)
    df$alc_with_meals.0.0[df$alc_freq.0.0 == "Never"] <- "None"
  }
  
  return(df)
}

ukb_train <- process_diet_none(ukb_train)
ukb_test  <- process_diet_none(ukb_test)
```



### SES skip logic

```{r, ses-skip-logic}
clean_employment_data <- function(ukb_df) {
  
  # employment indicator
  ukb_df <- ukb_df %>%
    mutate(
      employed = ifelse(
        ses_current_employment_status.0.6 %in% c("In paid employment or self-employed"),
        1, 0
      )
    )
  
  # convert to character
  char_vars <- c(
    "ses_job_shift_work.0.0",
    "ses_job_mainly_walking_or_standing.0.0",
    "ses_job_heavy_manual_or_physical_work.0.0",
    "ses_commute_transport_type.0.3",
    "ses_accom_type.0.0",
    "ses_household_income_before_tax_avg.0.0"
  )
  
  char_vars <- intersect(char_vars, names(ukb_df))
  ukb_df[char_vars] <- lapply(ukb_df[char_vars], as.character)
  
  # set employment dependent vars
  ukb_df$ses_length_working_week.0.0[ukb_df$employed == 0] <- 0
  ukb_df$ses_job_shift_work.0.0[ukb_df$employed == 0] <- "Not applicable"
  ukb_df$ses_job_mainly_walking_or_standing.0.0[ukb_df$employed == 0] <- "Not applicable"
  ukb_df$ses_job_heavy_manual_or_physical_work.0.0[ukb_df$employed == 0] <- "Not applicable"
  ukb_df$ses_commute_transport_type.0.3[ukb_df$employed == 0] <- "Not applicable"
  ukb_df$ses_commute_frequency.0.0[ukb_df$employed == 0] <- 0
  ukb_df$ses_commute_distance_home_to_work.0.0[ukb_df$employed == 0] <- 0
  
  
  ukb_df$ses_household_income_before_tax_avg.0.0[
    ukb_df$ses_accom_type.0.0 %in% c("Sheltered accommodation", "Care home")
  ] <- "Not applicable"
  
  # Return cleaned dataframe
  return(ukb_df)
}

ukb_train <- clean_employment_data(ukb_train)
ukb_test  <- clean_employment_data(ukb_test)
```

### Remove skip logic indicators

```{r, rm}

skip_indicators <- function(df){
  df %>%
    dplyr::select(-any_of(c(
      "employed",
      "smoke_past_indicator",
      "smoke_current_indicator",
      "ses_age_completed_full_time_education.0.0"
    )))
}

ukb_train <- skip_indicators(ukb_train)
ukb_test <- skip_indicators(ukb_test)

```

## 3) Table 1

```{r}
# ---- helper functions ----
fmt_mean_sd <- function(x, digits = 1) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  paste0(round(mean(x), digits), " (", round(sd(x), digits), ")")
}

fmt_n <- function(x) {
  sum(!is.na(x))
}

cat_table <- function(x, varname, digits = 1, include_na = FALSE) {
  if (!include_na) {
    tab <- table(x, useNA = "no")
    denom <- sum(tab)
  } else {
    tab <- table(x, useNA = "ifany")
    denom <- sum(tab)
  }
  prop <- (tab / denom) * 100
  
  data.frame(
    Variable = varname,
    Level    = names(tab),
    Summary  = paste0(as.integer(tab), " (", round(as.numeric(prop), digits), "%)"),
    stringsAsFactors = FALSE
  )
}

# ---- specify variables ----
cont_vars <- c("age_recr.0.0", "bmi.0.0", "sbp_avg")

cat_vars <- c("sex.0.0",
              "ethnicity.0.0",
              "cvd_event",
              "smoke_status.0.0",
              "alc_status.0.0",
              "ses_current_employment_status.0.6",
              "imd",
              "qualification.0.5")

# ---- build table 1 ----
table1_list <- list()

# add overall N
table1_list[["N_total"]] <- data.frame(
  Variable = "Participants",
  Level = "",
  Summary = as.character(nrow(ukb_train)),
  stringsAsFactors = FALSE
)

# continuous vars: Mean (SD)
for (v in cont_vars) {
  x <- ukb_train[[v]]
  table1_list[[v]] <- data.frame(
    Variable = v,
    Level = "",
    Summary = paste0(fmt_mean_sd(x), "; n=", fmt_n(x)),
    stringsAsFactors = FALSE
  )
}

# categorical vars: n (%)
for (v in cat_vars) {
  x <- ukb_train[[v]]
  table1_list[[v]] <- cat_table(x, varname = v, digits = 1, include_na = FALSE)
}

table1_df <- do.call(rbind, table1_list)

# view in console
print(table1_df, row.names = FALSE)

# export
write.csv(table1_df, file.path(output_dir, "Table1_imputed.csv"), row.names = FALSE)
```

## 4) Manhattan plot

```{r}
outcome <- "cvd_event"
covars  <- c("age_recr.0.0", "sex.0.0", "ethnicity.0.0",'bmi.0.0','sbp_avg','dbp_avg')
stopifnot(outcome %in% names(ukb_train), all(covars %in% names(ukb_train)))
ukb_train[[outcome]] <- as.integer(ukb_train[[outcome]])


# Define exposome exposures
drop_cols <- c(
  "eid","cvd_date","date_death.0.0","date_lost_to_fup.0.0","date_recr.0.0",
  "dob.0.0","mob","yob","assessment_centre_location.0.0",
  outcome, covars
)

exposome_prefix <- c("^diet_", "^emp_", "^env_", "^mh_", "^phys_", "^sed_", "^ses_", "^sleep_", "^smoke_")
exposome_single <- c(
  "qualification.0.5","alcohol_category"
)

exposures <- names(ukb_train)[
  Reduce(`|`, lapply(exposome_prefix, \(p) grepl(p, names(ukb_train)))) |
    names(ukb_train) %in% exposome_single
]
exposures <- setdiff(exposures, drop_cols)

# Domain mapping

get_domain <- function(v){
  if (grepl("^diet_|^alcohol_category", v)) return("Diet/Alcohol")
  if (grepl("^env_", v)) return("Environment")
  if (grepl("^mh_", v)) return("Mental health")
  if (grepl("^phys_|^sed_", v)) return("Physical activity")
  if (grepl("^ses_|^emp_|^imd$|^qualification|^employed$", v)) return("Socioeconomic")
  if (grepl("^sleep_", v)) return("Sleep")
  if (grepl("^smoke_", v)) return("Smoking")
  "Other"
}

# One-exposure EWAS model with GLOBAL p

run_one <- function(var) {
  dat <- ukb_train[, c(outcome, covars, var), drop = FALSE]
  if (is.character(dat[[var]])) dat[[var]] <- as.factor(dat[[var]])
  dat <- tidyr::drop_na(dat)
  
  # skip small sample
  n_used <- nrow(dat)
  if (n_used < 200) return(NULL)
  
  # skip constant
  if (length(unique(dat[[var]])) < 2) return(NULL)
  
  # logistic regression
  base_fml <- as.formula(paste0(outcome, " ~ ", paste(covars, collapse = " + ")))
  full_fml <- as.formula(paste0(outcome, " ~ ", var, " + ", paste(covars, collapse = " + ")))
  fit0 <- try(glm(base_fml, data = dat, family = binomial()), silent = TRUE)
  fit1 <- try(glm(full_fml, data = dat, family = binomial()), silent = TRUE)
  if (inherits(fit0, "try-error") || inherits(fit1, "try-error")) return(NULL)
  
  # exposure type
  is_fact <- is.factor(dat[[var]])
  
if (!is_fact) {
  # standardise continuous exposure if not binary
  if (length(unique(dat[[var]])) > 2) {
    dat[[var]] <- as.numeric(scale(dat[[var]]))
    type_now <- "numeric_sd"
  } else {
    type_now <- "numeric_binary"
  }
  
  fit1 <- try(glm(full_fml, data = dat, family = binomial()), silent = TRUE)
  if (inherits(fit1, "try-error")) return(NULL)
  
  sm <- summary(fit1)
  if (!(var %in% rownames(sm$coefficients))) return(NULL)
  
  beta <- sm$coefficients[var, "Estimate"]
  pval <- sm$coefficients[var, "Pr(>|z|)"]
  
  return(data.frame(
    variable = var,
    domain = get_domain(var),
    type = type_now,
    n_used = n_used,
    beta = beta,
    OR = exp(beta),
    p = pval,
    stringsAsFactors = FALSE
  ))
} else {
    # factor: GLOBAL LRT p
    lr <- anova(fit0, fit1, test = "LRT")
    pval <- lr$`Pr(>Chi)`[2]
    if (is.na(pval)) return(NULL)
    
    return(data.frame(
      variable = var,
      domain = get_domain(var),
      type = "factor_global",
      n_used = n_used,
      beta = NA_real_,
      OR = NA_real_,
      p = pval,
      stringsAsFactors = FALSE
    ))
  }
}

res <- lapply(exposures, run_one) %>% bind_rows()

res <- res %>%
  mutate(
    FDR = p.adjust(p, method = "BH"),
    neglog10p = -log10(p)
  )

write.csv(res, file.path(output_dir, "manhattan_results.csv"), row.names = FALSE)

# 4a) Plot Manhattan

domain_order <- c(
  "Diet/Alcohol",
  "Environment",
  "Mental health",
  "Physical activity",
  "Socioeconomic",
  "Sleep",
  "Smoking",
  "Other"
)

res2 <- res %>%
  mutate(domain = factor(domain, levels = domain_order)) %>%
  arrange(domain, variable) %>%
  mutate(pos = row_number())

axis_df <- res2 %>%
  group_by(domain) %>%
  summarise(center = (min(pos) + max(pos)) / 2, .groups = "drop")

bonf_thr <- 0.05 / nrow(res2)

p_manhattan <- ggplot(res2, aes(x = pos, y = neglog10p, color = domain)) +
  geom_point(size = 1.3, alpha = 0.85) +
  geom_hline(yintercept = -log10(bonf_thr), linetype = "dashed") +
  scale_x_continuous(breaks = axis_df$center, labels = axis_df$domain) +
  labs(
    title = "Adjusted Exposome Manhattan Plot (Train)",
    x = "Exposure domains",
    y = expression(-log[10](p))
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 25, hjust = 1)
  )

print(p_manhattan)
ggsave(file.path(output_dir, "manhattan_train_by_domain.png"), p_manhattan, width = 12, height = 5, dpi = 300)

## 4b) Top signals effect size table

# choose top variables from Manhattan results
top_vars <- res %>%
  arrange(p) %>%
  slice_head(n = 10) %>%
  pull(variable)

get_or_ci_table <- function(var) {
  dat <- ukb_train[, c(outcome, covars, var), drop = FALSE]
  dat <- tidyr::drop_na(dat)
  
  if (nrow(dat) < 200) return(NULL)
  if (length(unique(dat[[var]])) < 2) return(NULL)
  
  # character -> factor
  if (is.character(dat[[var]])) dat[[var]] <- as.factor(dat[[var]])
  
  # numeric: standardise to per SD increase
  if (is.numeric(dat[[var]]) && length(unique(dat[[var]])) > 2) {
    dat[[var]] <- as.numeric(scale(dat[[var]]))
    
    fit <- try(glm(
      as.formula(paste0(outcome, " ~ ", var, " + ", paste(covars, collapse = " + "))),
      data = dat,
      family = binomial()
    ), silent = TRUE)
    
    if (inherits(fit, "try-error")) return(NULL)
    
    sm <- summary(fit)$coefficients
    if (!(var %in% rownames(sm))) return(NULL)
    
    ci <- suppressMessages(confint.default(fit))
    
    data.frame(
      Variable = var,
      Level = "Per SD increase",
      OR = exp(sm[var, "Estimate"]),
      CI_low = exp(ci[var, 1]),
      CI_high = exp(ci[var, 2]),
      P_value = sm[var, "Pr(>|z|)"],
      stringsAsFactors = FALSE
    )
    
  } else if (is.factor(dat[[var]]) || is.logical(dat[[var]]) || (is.numeric(dat[[var]]) && length(unique(dat[[var]])) == 2)) {
    
    # if binary numeric, convert to factor for clearer output
    if (is.numeric(dat[[var]]) && length(unique(dat[[var]])) == 2) {
      dat[[var]] <- as.factor(dat[[var]])
    }
    
    fit <- try(glm(
      as.formula(paste0(outcome, " ~ ", var, " + ", paste(covars, collapse = " + "))),
      data = dat,
      family = binomial()
    ), silent = TRUE)
    
    if (inherits(fit, "try-error")) return(NULL)
    
    tidy_fit <- broom::tidy(fit, conf.int = TRUE, exponentiate = TRUE)
    
    # keep only exposure rows, not covariates / intercept
    out <- tidy_fit %>%
      dplyr::filter(grepl(paste0("^", var), term)) %>%
      dplyr::mutate(
        Variable = var,
        Level = sub(paste0("^", var), "", term),
        Level = ifelse(Level == "", "Per unit increase", Level),
        OR = estimate,
        CI_low = conf.low,
        CI_high = conf.high,
        P_value = p.value
      ) %>%
      dplyr::select(Variable, Level, OR, CI_low, CI_high, P_value)
    
    if (nrow(out) == 0) return(NULL)
    out
    
  } else {
    return(NULL)
  }
}

effect_table <- lapply(top_vars, get_or_ci_table) %>%
  bind_rows() %>%
  mutate(
    `OR (95% CI)` = sprintf("%.2f (%.2f–%.2f)", OR, CI_low, CI_high),
    P_value = sprintf("%.4f", P_value)
  ) %>%
  dplyr::select(Variable, Level, `OR (95% CI)`, P_value)

print(effect_table, row.names = FALSE)

write.csv(
  effect_table,
  file.path(output_dir, "top10_effect_size_table.csv"),
  row.names = FALSE
)

```

## 5) External exposome correlation analysis

Completing correlation analysis of the external exposome on train dataset before edits to check "raw" correlations between vars

#### Correlation analysis heatmap

```{r}

# Association functions 

cramers_v <- function(x, y){
  tab <- table(x,y)
  if(nrow(tab) < 2 || ncol(tab) < 2) return(NA_real_)
  chi <- suppressWarnings(chisq.test(tab, correct=FALSE))
  n <- sum(tab)
  k <- min(nrow(tab), ncol(tab))
  sqrt(as.numeric(chi$statistic)/(n*(k-1)))
}

eta_correlation <- function(y, x){
  x <- as.factor(x)
  if(!is.numeric(y) || sd(y, na.rm=TRUE)==0 || length(unique(x[!is.na(x)]))<=1) return(NA_real_)
  grand_mean <- mean(y, na.rm=TRUE)
  group_stats <- split(y,x)
  group_stats <- group_stats[lengths(group_stats) > 0]
  ss_between <- sum(sapply(group_stats, function(g) length(g)*(mean(g,na.rm=TRUE)-grand_mean)^2))
  ss_total <- sum((y-grand_mean)^2, na.rm=TRUE)
  if(ss_total==0) return(NA_real_)
  sqrt(ss_between/ss_total)
}

association_matrix <- function(df){
  num_vars <- names(df)[sapply(df,is.numeric)]
  cat_vars <- names(df)[sapply(df,is.factor)]
  
  # Numeric correlations
  cor_num <- if(length(num_vars) > 1) cor(df[num_vars], use="pairwise.complete.obs") else NULL
  
  # Categorical correlations
  cramer_matrix <- NULL
  if(length(cat_vars) > 1){
    cramer_matrix <- matrix(NA_real_, length(cat_vars), length(cat_vars), dimnames=list(cat_vars, cat_vars))
    for(i in seq_along(cat_vars)){
      for(j in i:length(cat_vars)){
        val <- if(i==j) 1 else cramers_v(df[[cat_vars[i]]], df[[cat_vars[j]]])
        cramer_matrix[i,j] <- val
        cramer_matrix[j,i] <- val
      }
    }
  }
  
  # Numeric vs categorical correlations
  num_cat_matrix <- NULL
  if(length(num_vars) > 0 && length(cat_vars) > 0){
    num_cat_matrix <- matrix(NA_real_, length(num_vars), length(cat_vars), dimnames=list(num_vars, cat_vars))
    for(i in seq_along(num_vars)){
      for(j in seq_along(cat_vars)){
        num_cat_matrix[i,j] <- eta_correlation(df[[num_vars[i]]], df[[cat_vars[j]]])
      }
    }
  }
  
  # Combine matrices
  top <- if(!is.null(cor_num) && !is.null(num_cat_matrix)) cbind(cor_num, num_cat_matrix) else cor_num
  bottom <- if(!is.null(cramer_matrix) && !is.null(num_cat_matrix)) cbind(t(num_cat_matrix), cramer_matrix) else cramer_matrix
  assoc_mat <- if(!is.null(top) && !is.null(bottom)) rbind(top, bottom) else if(!is.null(top)) top else bottom
  
  # Remove invalid variables
  if(!is.null(assoc_mat)){
    bad_vars <- rownames(assoc_mat)[apply(assoc_mat, 1, function(x) any(is.na(x) | is.nan(x) | is.infinite(x)))]
    if(length(bad_vars) > 0){
      assoc_mat <- assoc_mat[!rownames(assoc_mat) %in% bad_vars, !colnames(assoc_mat) %in% bad_vars, drop=FALSE]
    }
    if(nrow(assoc_mat) == ncol(assoc_mat) && nrow(assoc_mat) > 0) diag(assoc_mat) <- 1
  }
  
  return(as.matrix(assoc_mat))
}

# Prepare combined dataframe 

domain_list <- list(
  lifestyle = ukb_train %>% dplyr::select(matches("^(alc|diet|phys|sed|smoke|sleep)")),
  socioeconomic = ukb_train %>% dplyr::select(matches("^(ses|emp)"), qualification.0.5, imd, ethnicity.0.0),
  psychosocial = ukb_train %>% dplyr::select(matches("^mh")),
  environmental = ukb_train %>% dplyr::select(matches("^env"))
)

# Convert characters to factors
domain_list <- lapply(domain_list, function(df){
  df[] <- lapply(df, function(x) if(is.character(x)) as.factor(x) else x)
  df
})

# Combine domains
big_df <- as.data.frame(bind_cols(domain_list))

# Compute association matrix
assoc_big <- association_matrix(big_df)

# Plot combined heatmap 
png(file.path(output_dir, "all_domains_association_heatmap.png"), width=4000, height=4000, res=300)
corrplot(
  assoc_big,
  method="color",
  type="upper",
  order="hclust",
  col=colorRampPalette(c("#3B4CC0","white","#B40426"))(200),
  tl.cex=0.6,
  tl.col="black",
  cl.cex=0.8,
  diag=FALSE
)

title(main="All domains mixed-type association heatmap")
dev.off()

# Convert association matrix to long format
assoc_long <- as.data.frame(as.table(assoc_big))
names(assoc_long) <- c("Variable1", "Variable2", "Association")

# Remove self-correlations and duplicates
assoc_long <- assoc_long %>%
  dplyr::filter(Variable1 != Variable2) %>%
  dplyr::filter(as.character(Variable1) < as.character(Variable2))

# Define thresholds
thresholds <- c(0.5, 0.7, 0.9)

# Loop over thresholds and save CSVs
for(thresh in thresholds){
  strong_df <- assoc_long %>%
    dplyr::filter(abs(Association) >= thresh) %>%
    arrange(desc(abs(Association)))
  
  # Print top 30 to console
  cat(paste0("\nAssociations ≥ ", thresh, "\n"))
  print(head(strong_df, 30))
  
  # Save to CSV
  write.csv(
    strong_df,
    file = file.path(output_dir, paste0("strong_associations_", gsub("\\.", "", thresh*100), ".csv")),
    row.names = FALSE
  )
}

```

#### Baseline vars analysis

```{r, baseline}

# analysing baseline vars relationship

# Subset data for variables of interest
plot_df <- ukb_train %>%
  dplyr::select(age_recr.0.0, bmi.0.0, sbp_avg, dbp_avg, sex.0.0) %>%  # adjust variable names if needed
  dplyr::filter(!is.na(age_recr.0.0) & !is.na(bmi.0.0) & !is.na(sbp_avg) & !is.na(dbp_avg) & !is.na(sex.0.0))

# Ensure Sex is factor
plot_df$sex.0.0 <- as.factor(plot_df$sex.0.0)


# Create pair plot (scatterplot matrix with densities)

pair_plot <- ggpairs(
  plot_df,
  columns = 1:4,          # numeric columns for scatterplot matrix
  aes(color = sex.0.0, alpha = 0.4),
  upper = list(continuous = wrap("points", size = 0.5)),
  lower = list(continuous = wrap("points", size = 0.5)),
  diag = list(continuous = wrap("densityDiag", alpha = 0.5))
) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 10),
    axis.text = element_text(size = 8),
    legend.position = "bottom"
  )


# Save to file
ggsave(file.path(output_dir, "age_bmi_sbp_dbp_pairs.png"), pair_plot, width = 10, height = 10, dpi = 300)

# Display plot
pair_plot

```

#### Edit dataset to deal with high colinearity (\>=0.75)

```{r}
strong_075 <- assoc_long %>%
  dplyr::filter(abs(Association)>=0.75) %>%
  arrange(desc(abs(Association)))

cat("\nAssociations ≥ 0.75\n")
print(head(strong_075,30))

```

## 6) One hot encoding & deterministic recoding of categorical vars

#### Mh_activities

one hot encoding

```{r, mh_activities}

train_v1 <- ukb_train
test_v1 <- ukb_test


process_mh_activities <- function(df) {
  
  # Identify target columns
  mh_cols_target <- paste0("mh_activities.0.", 0:4)
  mh_cols_present <- mh_cols_target[mh_cols_target %in% names(df)]
  
  if (length(mh_cols_present) == 0) {
    stop("No mh_activities columns found.")
  }
  
  # Convert to character matrix
  mh_mat_chr <- as.matrix(
    df[, mh_cols_present, drop = FALSE] %>%
      mutate(across(everything(), as.character))
  )
  
  # Collect unique codes
  all_codes <- sort(unique(c(mh_mat_chr)))
  all_codes <- all_codes[!is.na(all_codes)]
  
  # Safe column name helper
  safe_code_name <- function(x) {
    x <- as.character(x)
    x <- tolower(x)
    x <- gsub("[^A-Za-z0-9]+", "_", x)
    x <- gsub("^_+|_+$", "", x)
    x
  }
  
  onehot_names <- paste0("mh_activities_", safe_code_name(all_codes))
  
  # Build one-hot matrix
  onehot <- matrix(
    0L,
    nrow = nrow(df),
    ncol = length(all_codes),
    dimnames = list(NULL, onehot_names)
  )
  
  for (j in seq_along(all_codes)) {
    code <- all_codes[j]
    onehot[, j] <- as.integer(rowSums(mh_mat_chr == code, na.rm = TRUE) > 0)
  }
  
  # Bind to df
  df <- bind_cols(df, as.data.frame(onehot))
  
  # Drop original array columns
  df <- df %>%
    dplyr::select(-all_of(mh_cols_present))
  
  return(df)
}

train_v1 <- process_mh_activities(train_v1)
test_v1  <- process_mh_activities(test_v1)

```

deterministic recoding

```{r, mh_activities_dr}
mh_exclusive <- function(df) {
  
  # Identify mh one-hot columns
  mh_cols <- grep("^mh_activities_", names(df), value = TRUE)
  
  none_col <- "mh_activities_none_of_the_above"
  pna_col  <- "mh_activities_prefer_not_to_answer"
  
  # Only operate on columns that actually exist
  existing_special <- intersect(c(none_col, pna_col), names(df))
  other_cols <- setdiff(mh_cols, existing_special)
  
  # Apply exclusivity safely
  if (none_col %in% names(df)) {
    idx_none <- df[[none_col]] == 1
    df[idx_none, other_cols] <- 0
  }
  
  if (pna_col %in% names(df)) {
    idx_pna <- df[[pna_col]] == 1
    df[idx_pna, other_cols] <- 0
  }
  
  return(df)
}

train_v1 <- mh_exclusive(train_v1)
test_v1  <- mh_exclusive(test_v1)
```


## 7) Score creation
### Diet score

```{r, diet}
process_diet <- function(df) {
  
  # Ensure we are working with data.table
  if (!"data.table" %in% class(df)) df <- as.data.table(df)
  
  vars_to_drop_1 <- paste0(c(
  "diet_spread_type",
  "diet_salt_added_food",
  "diet_milk_type",
  "diet_processed_meat_intake",
  "diet_pork_intake",
  "diet_oily_fish_intake",
  "diet_non_oily_fish_intake",
  "diet_lamb_mutton_intake",
  "diet_cereal_intake",
  "diet_beef_intake",
  "diet_fresh_fruit_intake",
  "diet_dried_fruit_intake",
  "diet_cooked_veg_intake",
  "diet_raw_veg_intake",
  "diet_water_intake"
), ".0.0")
  
  # Convert fruit/veg to numeric
  fv_to_num <- function(x) {
    x <- trimws(as.character(x))
    x[x %in% c("Less than once", "Less than one")] <- "0"
    suppressWarnings(as.numeric(x))
  }
  
  freq_to_week <- function(x) {
    x <- as.character(x)
    out <- rep(NA_real_, length(x))
    out[x == "Never"] <- 0
    out[x == "Less than once a week"] <- 0.5
    out[x == "Once a week"] <- 1
    out[x == "2-4 times a week"] <- 3
    out[x == "5-6 times a week"] <- 5.5
    out[x == "Once or more daily"] <- 7
    out
  }
  
  df[, fv_fresh  := fv_to_num(diet_fresh_fruit_intake.0.0)]
  df[, fv_dried  := fv_to_num(diet_dried_fruit_intake.0.0)]
  df[, fv_cooked := fv_to_num(diet_cooked_veg_intake.0.0)]
  df[, fv_raw    := fv_to_num(diet_raw_veg_intake.0.0)]
  
  # Totals
  df[, diet_total_fish_intake :=
       freq_to_week(diet_oily_fish_intake.0.0) +
       freq_to_week(diet_non_oily_fish_intake.0.0)]
  
  df[, diet_total_red_meat_intake :=
       freq_to_week(diet_lamb_mutton_intake.0.0) +
       freq_to_week(diet_beef_intake.0.0) +
       freq_to_week(diet_pork_intake.0.0)]
  
  df[, diet_total_fruit_veg_intake := fv_fresh + fv_dried + fv_cooked + fv_raw]
  
  # Dichotomise 
  df[, risk_fruit_veg :=
       fifelse(diet_total_fruit_veg_intake >= 5, 0, 1)]
  
  df[, risk_fish :=
       fifelse(diet_total_fish_intake >= 2, 0, 1)]
  
  df[, risk_processed_meat :=
       fifelse(freq_to_week(diet_processed_meat_intake.0.0) <= 1, 0, 1)]
  
  df[, risk_red_meat :=
       fifelse(diet_total_red_meat_intake <= 1, 0, 1)]
  
  df[, risk_milk :=
       fifelse(diet_milk_type.0.0 %in% c("Semi-skimmed", "Skimmed"), 0, 1)]
  
  df[, risk_spread :=
       fifelse(diet_spread_type.0.0 == "Never/rarely", 0, 1)]
  
  df[, risk_salt :=
       fifelse(diet_salt_added_food.0.0 == "Never/rarely", 0, 1)]
  
  df[, risk_cereal :=
       fifelse(as.numeric(diet_cereal_intake.0.0) > 5, 0, 1)]
  
  df[, risk_water :=
       fifelse(as.numeric(diet_water_intake.0.0) >= 6, 0, 1)]
  
  # Total dietary risk score 
  df[, diet_risk_score :=
       risk_fruit_veg +
       risk_fish +
       risk_processed_meat +
       risk_red_meat +
       risk_milk +
       risk_spread +
       risk_salt +
       risk_cereal +
       risk_water]
  
  df <- df %>%
  dplyr::select(
    -starts_with("risk_"),
    -starts_with("diet_total_f"),
    -starts_with("fv_"),
    -diet_total_red_meat_intake,
    -dplyr::any_of(vars_to_drop_1)
  )
  
  return(df)
}

train_v1 <- process_diet(train_v1)
test_v1  <- process_diet(test_v1)

```
### Sleep score

```{r, sleep}
# Use 5 sleep factors: 
# "sleep_daytime_dozing.0.0" <- "Sometimes", "Often", "All of the time" ~ 1, "Never/rarely" ~ 0
# "sleep_insomnia.0.0" <- "Sometimes", "Usually" ~ 1, "Never/rarely" ~ 0
# "sleep_snore.0.0" <- "Yes" ~ 1, "No" ~ 0 
# "sleep_chronotype.0.0" <- "More an 'evening' than a 'morning' person" | "Definitely an 'evening' person" ~ 1, "Definitely a 'morning' person" | "More a 'morning' than 'evening' person" ~ 0
# "sleep_duration.0.0" <- <7 | > 8 ~ 1, >=7 | <=8  ~ 0


create_sleep_score <- function(df) {
  
  df %>%
    
    # Ensure correct types first
    mutate(
      sleep_duration.0.0 = as.numeric(as.character(sleep_duration.0.0)),
      sleep_chronotype.0.0 = as.character(sleep_chronotype.0.0),
      sleep_insomnia.0.0 = as.character(sleep_insomnia.0.0),
      sleep_snore.0.0 = as.character(sleep_snore.0.0),
      sleep_daytime_dozing.0.0 = as.character(sleep_daytime_dozing.0.0)
    ) %>%
    
    mutate(
      
      # Sleep duration (<7 or >8 hours = unhealthy)
      sleep_duration_score = case_when(
        sleep_duration.0.0 < 7 ~ 1,
        sleep_duration.0.0 > 8 ~ 1,
        sleep_duration.0.0 >= 7 & sleep_duration.0.0 <= 8 ~ 0,
        TRUE ~ NA_real_
      ),
      
      # Chronotype
      sleep_chronotype_score = case_when(
        sleep_chronotype.0.0 %in% c(
          "More an 'evening' than a 'morning' person",
          "Definitely an 'evening' person"
        ) ~ 1,
        sleep_chronotype.0.0 %in% c(
          "Definitely a 'morning' person",
          "More a 'morning' than 'evening' person"
        ) ~ 0,
        TRUE ~ NA_real_
      ),
      
      # Insomnia
      sleep_insomnia_score = case_when(
        sleep_insomnia.0.0 %in% c("Sometimes", "Usually") ~ 1,
        sleep_insomnia.0.0 == "Never/rarely" ~ 0,
        TRUE ~ NA_real_
      ),
      
      # Snoring
      sleep_snore_score = case_when(
        sleep_snore.0.0 == "Yes" ~ 1,
        sleep_snore.0.0 == "No" ~ 0,
        TRUE ~ NA_real_
      ),
      
      # Daytime sleepiness
      sleep_daytime_score = case_when(
        sleep_daytime_dozing.0.0 %in% c("Sometimes", "Often", "All of the time") ~ 1,
        sleep_daytime_dozing.0.0 == "Never/rarely" ~ 0,
        TRUE ~ NA_real_
      ),
      
      # Total score
      sleep_score = sleep_duration_score +
        sleep_chronotype_score +
        sleep_insomnia_score +
        sleep_daytime_score +
        sleep_snore_score,
      
      # Category
      sleep_category = case_when(
        sleep_score %in% 0:1 ~ "Healthy",
        sleep_score %in% 2:3 ~ "Intermediate",
        sleep_score %in% 4:5 ~ "Poor",
        TRUE ~ NA_character_
      )
    ) %>%
    
    # Remove intermediate variables
    select(
      -sleep_chronotype.0.0,
      -sleep_daytime_dozing.0.0,
      -sleep_duration.0.0,
      -sleep_insomnia.0.0,
      -sleep_snore.0.0,
      -sleep_chronotype_score,
      -sleep_daytime_score,
      -sleep_duration_score,
      -sleep_insomnia_score,
      -sleep_snore_score,
      -sleep_score
    ) %>%
    
    # Ordered factor
    mutate(
      sleep_category = factor(
        sleep_category,
        levels = c("Healthy", "Intermediate", "Poor"),
        ordered = TRUE
      )
    )
} 

# Run on train and test set

train_v1 <- create_sleep_score(train_v1)
test_v1  <- create_sleep_score(test_v1)

```

### MET score

MET score calculated based on IPAQ

Walking MET-minutes/week = 3.3 \* walking minutes \* walking days Moderate MET-minutes/week = 4.0 \* moderate-intensity activity minutes \* moderate days Vigorous MET-minutes/week = 8.0 \* vigorous-intensity activity minutes \* vigorous-intensity days Total physical activity MET-minutes/week = sum of Walking + Moderate + Vigorous METminutes/week scores.

```{r}
### Physical activity MET score

create_met_total <- function(df) {
  
  met_vars <- c(
    "phys_MET_walking.0.0",
    "phys_MET_moderate_activity.0.0",
    "phys_MET_vigorous_activity.0.0"
  )
  
  # ensure numeric
  # df[met_vars] <- lapply(df[met_vars], function(x) as.numeric(as.character(x)))
  
  # truncate impossible negative values to 0
  for (v in met_vars) {
    df[[v]] <- pmax(df[[v]], 0)
  }
  
  # total MET
  df <- df %>%
    mutate(
      phys_MET_total =
        phys_MET_walking.0.0 +
        phys_MET_moderate_activity.0.0 +
        phys_MET_vigorous_activity.0.0
    )
  
  return(df)
}

train_v1 <- create_met_total(train_v1)
test_v1  <- create_met_total(test_v1)
```

### MH Symptom score

```{r}
### Mental health symptom score

create_mh_symptom_score <- function(df) {
  
  score_map <- c(
    "Not at all" = 0,
    "Several days" = 1,
    "More than half the days" = 2,
    "Nearly every day" = 3
  )
  
  df <- df %>%
    dplyr::mutate(
      mh_depression_score   = unname(score_map[as.character(mh_frequency_depression_2_weeks.0.0)]),
      mh_tenseness_score    = unname(score_map[as.character(mh_frequency_tenseness_2_weeks.0.0)]),
      mh_tiredness_score    = unname(score_map[as.character(mh_frequency_tiredness_2_weeks.0.0)]),
      mh_unenthusiasm_score = unname(score_map[as.character(mh_frequency_unenthusiasm_2_weeks.0.0)]),
      
      mh_symptom_score =
        mh_depression_score +
        mh_tenseness_score +
        mh_tiredness_score +
        mh_unenthusiasm_score
    ) %>%
    dplyr::select(
      -mh_frequency_depression_2_weeks.0.0,
      -mh_frequency_tenseness_2_weeks.0.0,
      -mh_frequency_tiredness_2_weeks.0.0,
      -mh_frequency_unenthusiasm_2_weeks.0.0,
      -mh_depression_score,
      -mh_tenseness_score,
      -mh_tiredness_score,
      -mh_unenthusiasm_score
    )
  
  return(df)
}

train_v1 <- create_mh_symptom_score(train_v1)
test_v1  <- create_mh_symptom_score(test_v1)
```

## 8) Categories

### Alcohol and BMI cats

```{r, alc_bmi}

process_alcohol_bmi <- function(df) {
  
  df %>%
    dplyr::mutate(
      
      #aclohol categories
      alcohol_category = case_when(
        
        # Never drinkers
        alc_freq.0.0 == "Never" ~ "Non-drinker",
        alc_status.0.0 == "Never" ~ "Non-drinker",
        
        # Former drinkers
        alc_status.0.0 == "Previous" ~ "Former",
        
        # Current drinkers by frequency
        alc_status.0.0 == "Current" &
          alc_freq.0.0 %in% c("Special occasions only",
                              "One to three times a month") ~ "Social drinker",
        
        alc_status.0.0 == "Current" &
          alc_freq.0.0 == "Once or twice a week" ~ "Moderate drinker",
        
        alc_status.0.0 == "Current" &
          alc_freq.0.0 %in% c("Three or four times a week",
                              "Daily or almost daily") ~ "Daily drinker",
        
        TRUE ~ NA_character_
      ),
      
      # bmi cats
      bmi = cut(
        bmi.0.0,
        breaks = c(-Inf, 18.5, 25, 30, 40, Inf),
        labels = c(
          "Underweight (<18.5)",
          "Normal weight (18.5–24.9)",
          "Overweight (25–29.9)",
          "Obese (30–39.9)",
          "Severely obese (≥40)"
        ),
        right = FALSE
      )
    ) %>%
    
# drop original cols
  dplyr::select(-bmi.0.0, -alc_freq.0.0, -alc_status.0.0,)
}  
train_v1 <- process_alcohol_bmi(train_v1)
test_v1  <- process_alcohol_bmi(test_v1)
```

### Employment cats & one hot encoding

```{r, employment}
process_employment <- function(df) {
  
  emp_col <- "ses_current_employment_status.0.6"
  
  if (!emp_col %in% names(df)) {
    stop("Employment column not found.")
  }
  
  # Ensure character
  emp_chr <- as.character(df[[emp_col]])
  
  # Collapse categories
  employment_collapsed <- case_when(
    emp_chr == "In paid employment or self-employed" ~ 
      "in_paid_employment_or_self_employed",
    
    emp_chr == "Retired" ~ 
      "retired",
    
    !is.na(emp_chr) ~ 
      "other",
    
    TRUE ~ NA_character_
  )
  
  df$employment_collapsed <- employment_collapsed
  
  # One hot encoding
  categories <- c(
    "in_paid_employment_or_self_employed",
    "retired",
    "other"
  )
  
  onehot <- sapply(categories, function(cat) {
    as.integer(df$employment_collapsed == cat)
  })
  
  colnames(onehot) <- paste0("emp_", categories)
  
  df <- bind_cols(df, as.data.frame(onehot))
  
  df <- df%>%
    dplyr::select(-employment_collapsed, -ses_current_employment_status.0.6)
  
  return(df)
}

train_v1 <- process_employment(train_v1)
test_v1  <- process_employment(test_v1)

```

### Reorder

```{r}
reorder_cols <- function(df){

  first_cols <- c("eid", "cvd_event", "cvd_date")

  other_cols <- setdiff(names(df), first_cols)

  df %>%
    dplyr::select(all_of(first_cols), all_of(sort(other_cols)))

}

train_v1 <- reorder_cols(train_v1)
test_v1  <- reorder_cols(test_v1)
```

## 9) External exposome specific dataset

```{r}
outcome <- "cvd_event"
covars  <- c("age_recr.0.0", "sex.0.0", "ethnicity.0.0",'bmi','sbp_avg','dbp_avg')
drop_cols <- c(
  "eid","cvd_date","date_death.0.0","date_lost_to_fup.0.0","date_recr.0.0",
  "dob.0.0","mob","yob","assessment_centre_location.0.0",
  outcome, covars
)

exposome_prefix <- c("^diet_", "^emp_", "^env_", "^mh_", "^phys_", "^sed_", "^ses_", "^sleep_", "^smoke_")
exposome_single <- c(
  "qualification.0.5","alcohol_category","employed"
)

exposures <- names(train_v1)[
  Reduce(`|`, lapply(exposome_prefix, \(p) grepl(p, names(train_v1)))) |
    names(train_v1) %in% exposome_single
]
exposures <- setdiff(exposures, drop_cols)

```

##### Save

```{r, save}
saveRDS(train_v1, file = file.path(output_dir, "train_final.rds"))
saveRDS(test_v1,  file = file.path(output_dir, "test_final.rds"))
```
